```{r setup}
source("Rscripts/utils.R")

# Haplotype category colors
shared_color <- "#888888"
hap_distinct_color <- "#8b0000"

hdca_colors <- c(
    "Shared" = shared_color,
    "Haplotype-distinct" = hap_distinct_color,
    "Epigenetically haplotype-selective" = "purple",
    "Genetically haplotype-specific" = "#f0a500"
)
name_order <- rev(names(hdca_colors))
```

# Load tables (produced by union-peaks.Rmd)
```{r load-tables}
union_df <- fread("Tables/union-peaks.tsv.gz")
union_df[, haplotype_distinct := factor(haplotype_distinct, levels = name_order)]

hap_wide <- fread("Tables/union-peaks-hap-wide.tsv.gz")
hap_wide[, haplotype_distinct := factor(haplotype_distinct, levels = name_order)]
```

# Peaks per sample
```{r peaks-per-sample}
union_df %>%
    group_by(Individual_ID, PS, Platform, Haplotype, sample_id) %>%
    summarise(n_peaks = sum(is_peak)) %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = n_peaks, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("# of peaks per sample", labels = comma) +
    xlab("") +
    my_grid() +
    scale_color_manual(values = PLAT_COLORS) +
    theme(legend.position = "top")

my_ggsave("Figures/union-peaks-per-sample.pdf", width = 3, height = 3)
```

# Peak sharing across samples
```{r peak-sharing}
peak_sharing <- union_df[,
    .(
        n_samples_peak = sum(is_peak),
        n_samples_total = .N,
        exists_in_hg38 = exists_in_hg38[1]
    ),
    by = consensus_peak_id
]

peak_sharing %>%
    ggplot(aes(x = n_samples_peak)) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "black", linewidth = 0.2) +
    scale_x_continuous("# samples where site is a peak",
        breaks = scales::breaks_width(1)
    ) +
    scale_y_continuous("# consensus peaks", labels = comma) +
    my_grid()

my_ggsave("Figures/union-peak-sharing-histogram.pdf", width = 5, height = 3)
```

# Peak sharing stratified by hg38 liftover
```{r peak-sharing-hg38}
peak_sharing %>%
    ggplot(aes(x = n_samples_peak, fill = exists_in_hg38)) +
    geom_histogram(binwidth = 1, position = "dodge", color = "black", linewidth = 0.2) +
    scale_x_continuous("# samples where site is a peak",
        breaks = scales::breaks_width(1)
    ) +
    scale_y_continuous("# consensus peaks", labels = comma) +
    scale_fill_brewer("Exists in hg38", palette = "Set1") +
    my_grid() +
    theme(legend.position = "top")

my_ggsave("Figures/union-peak-sharing-by-hg38.pdf", width = 5, height = 3)
```

# Accessibility distribution
```{r accessibility-distribution}
set.seed(42)
plot_sample <- union_df[sample(.N, min(.N, 1e6))]

plot_sample %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = is_peak)) +
    geom_histogram(binwidth = 0.02, position = "identity", alpha = 0.6) +
    geom_vline(xintercept = 0.20, linetype = "dashed", color = "red") +
    scale_x_continuous("FIRE coverage / total coverage", limits = c(0, 1)) +
    scale_y_continuous("# sites (sampled)", labels = comma) +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray70")) +
    my_grid() +
    theme(legend.position = "top")

my_ggsave("Figures/union-accessibility-distribution.pdf", width = 5, height = 3)
```

# Accessibility by platform
```{r accessibility-by-platform}
plot_sample %>%
    filter(is_peak, !is.na(Platform)) %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = Platform)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = 0.20, linetype = "dashed", color = "red") +
    scale_x_continuous("FIRE coverage / total coverage", limits = c(0, 1)) +
    scale_fill_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")

my_ggsave("Figures/union-accessibility-by-platform.pdf", width = 5, height = 3)
```

# Original peak recovery
```{r original-peak-recovery}
orig_recovery <- union_df[is_peak == TRUE & !is.na(Platform),
    .(
        n_peaks = .N,
        n_overlap_orig = sum(overlaps_original_called_peak == TRUE),
        frac_overlap_orig = mean(overlaps_original_called_peak == TRUE)
    ),
    by = .(Individual_ID, Platform, Haplotype, sample_id)
]

orig_recovery %>%
    ggplot(aes(x = "", y = frac_overlap_orig * 100, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("% peaks overlapping original call") +
    xlab("") +
    my_grid() +
    scale_color_manual(values = PLAT_COLORS) +
    theme(legend.position = "top")

my_ggsave("Figures/union-original-peak-recovery.pdf", width = 3, height = 3)
```

# hg38 liftover rate
```{r hg38-liftover}
hg38_rate <- union_df[is_peak == TRUE & !is.na(Platform),
    .(
        n_peaks = .N,
        n_in_hg38 = sum(exists_in_hg38 == TRUE),
        frac_hg38 = mean(exists_in_hg38 == TRUE)
    ),
    by = .(Individual_ID, Platform, Haplotype, sample_id)
]

hg38_rate %>%
    ggplot(aes(x = "", y = frac_hg38 * 100, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("% peaks with hg38 coordinates") +
    xlab("") +
    my_grid() +
    scale_color_manual(values = PLAT_COLORS) +
    theme(legend.position = "top")

my_ggsave("Figures/union-hg38-liftover-rate.pdf", width = 3, height = 3)
```

# Peak sharing categories per sample
```{r peak-sharing-per-sample-view}
sharing_info <- peak_sharing[, .(consensus_peak_id, n_samples_peak)]

sample_sharing <- merge(
    union_df[is_peak == TRUE & !is.na(Platform),
             .(sample_id, consensus_peak_id, Platform)],
    sharing_info,
    by = "consensus_peak_id"
)

sample_sharing[, sharing_cat := fcase(
    n_samples_peak == 1, "Private (1)",
    n_samples_peak <= 5, "Rare (2-5)",
    n_samples_peak <= 20, "Low (6-20)",
    n_samples_peak <= 45, "Medium (21-45)",
    default = "High (46+)"
)]
sample_sharing[, sharing_cat := factor(sharing_cat,
    levels = c("Private (1)", "Rare (2-5)", "Low (6-20)",
               "Medium (21-45)", "High (46+)")
)]

sample_sharing[, .(n_peaks = .N), by = .(sample_id, Platform, sharing_cat)] %>%
    ggplot(aes(x = sharing_cat, y = n_peaks, color = Platform)) +
    geom_quasirandom(method = "pseudorandom", dodge.width = 0.8, size = 0.8) +
    scale_y_continuous("# peaks per sample", labels = comma) +
    xlab("Peak sharing category") +
    my_grid() +
    scale_color_manual(values = PLAT_COLORS) +
    theme(
        legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1),
    )

my_ggsave("Figures/union-peak-sharing-categories.pdf", width = 5, height = 3.5)
```

# ============================================================
# Haplotype analysis plots
# ============================================================

# Per-sample counts summary
```{r per-sample-summary}
per_sample_summary <- union_df[is_peak == TRUE, .(
    n = .N,
    n_autosome = sum(autosome == "Autosomes"),
    n_sex = sum(autosome == "Sex chromosome"),
    n_sd = sum(is_SD),
    percent_sd = 100 * sum(is_SD) / .N,
    mean_coverage = mean(coverage)
), by = .(Individual_ID, ind_key, Platform)]
per_sample_summary
```

# H1 vs H2 fire coverage scatter (HG002)
```{r hcsa-scatter}
hap_wide[Individual_ID == "HG002" & autosome == "Autosomes" &
         peak_status == "both"] %>%
    ggplot(aes(x = fire_coverage_1, y = fire_coverage_2)) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    scale_x_continuous("H1 FIRE coverage") +
    scale_y_continuous("H2 FIRE coverage") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    facet_col(~ind_key, scales = "free") +
    my_grid()
my_ggsave("Figures/union-hcsa-scatter.pdf", width = 4, height = 3)
```

# Category summary and bar plot (HG002)
```{r hap-distinct-bar}
hap_summary <- hap_wide[, .(n = .N),
    by = .(Individual_ID, ind_key, Platform, autosome, haplotype_distinct)
][, `:=`(
    total_n = sum(n),
    frac = n / sum(n)
), by = ind_key]

hap_summary[Individual_ID == "HG002" & autosome == "Autosomes" &
            haplotype_distinct != "Shared"] %>%
    ggplot(aes(y = haplotype_distinct, fill = haplotype_distinct, x = n)) +
    geom_bar(stat = "identity") +
    geom_text(
        data = . %>%
            group_by(ind_key, haplotype_distinct) %>%
            summarise(
                n = sum(n), frac = sum(frac),
                label = paste0(comma(n), " (", round(100 * frac, 1), "%)")
            ) %>%
            ungroup(),
        aes(x = 0, label = label, fill = NULL),
        hjust = -0.1, color = "black", size = 2
    ) +
    scale_fill_manual(values = hdca_colors) +
    scale_x_continuous("# of regulatory elements", label = comma) +
    ylab("") +
    facet_col(~ind_key) +
    my_grid() +
    theme(legend.position = "none")
my_ggsave("Figures/union-haplotype-distinct-bar.pdf", width = 3, height = 1.5)
```

# Pie chart: shared vs haplotype-distinct (HG002)
```{r hap-distinct-pie}
hap_summary[Individual_ID == "HG002" & autosome == "Autosomes"] %>%
    mutate(hap_cat = ifelse(
        haplotype_distinct == "Shared", "Shared", "Haplotype-distinct"
    )) %>%
    group_by(hap_cat, ind_key) %>%
    summarise(n = sum(n)) %>%
    ungroup() %>%
    mutate(
        percent = 100 * n / sum(n),
        label = paste0("n=", comma(n), "\n", round(percent, 1), "%"),
    ) %>%
    ggplot(aes(x = "", y = n, fill = hap_cat)) +
    geom_col() +
    coord_polar(theta = "y", start = pi / 2) +
    geom_text(
        aes(label = label),
        position = position_stack(vjust = 0.5),
        size = 1.5, color = "white"
    ) +
    scale_fill_manual("", values = hdca_colors) +
    guides(fill = guide_legend(ncol = 1)) +
    facet_col(~ind_key) +
    theme_void(base_size = 6) +
    theme(legend.position = "top", legend.key.size = unit(0.2, "cm"))
my_ggsave("Figures/union-haplotype-distinct-pie.pdf", width = 1.5, height = 1.5)
```

# Beeswarm: % haplotype-distinct across samples
```{r hap-distinct-beeswarm}
hap_summary[autosome == "Autosomes" & haplotype_distinct != "Shared",
    .(pct_hap_distinct = sum(frac)),
    by = ind_key
] %>%
    ggplot(aes(x = "", y = pct_hap_distinct)) +
    geom_violin(fill = "grey", color = NA, alpha = 0.5,
        adjust = 1.5, scale = "width", width = 0.3) +
    geom_quasirandom(
        width = 0.1, alpha = 0.5, size = 1,
        color = hap_distinct_color
    ) +
    scale_y_continuous("REs that are haplotype-distinct",
        limits = c(0, NA),
        labels = percent_format(accuracy = 1)) +
    xlab("HPRC Fiber-seq pilot samples") +
    my_grid()
my_ggsave("Figures/union-haplotype-distinct-beeswarm.pdf",
    width = 1.5, height = 2)
```

# % haplotype-distinct vs mean coverage
```{r hap-distinct-vs-coverage}
hap_distinct_frac <- hap_summary[autosome == "Autosomes",
    .(frac_hap_distinct = sum(frac[haplotype_distinct != "Shared"])),
    by = .(Individual_ID, ind_key, Platform)
]
coverage_info <- per_sample_summary[, .(ind_key, mean_coverage)]
hap_vs_cov <- merge(hap_distinct_frac, coverage_info, by = "ind_key")

hap_vs_cov %>%
    ggplot(aes(x = mean_coverage, y = frac_hap_distinct)) +
    stat_cor(method = "pearson", size = 2) +
    geom_point(size = 2, color = hap_distinct_color, alpha = 0.7) +
    scale_x_continuous("Mean coverage") +
    scale_y_continuous("REs that are haplotype-distinct", labels = percent) +
    my_grid()
my_ggsave("Figures/union-hap-distinct-vs-coverage.pdf", width = 2, height = 2)
```

# Total REs per sample
```{r total-res-per-sample}
hap_summary[autosome == "Autosomes",
    .(n = sum(n)),
    by = .(ind_key, Platform)
] %>%
    ggplot(aes(x = "", y = n)) +
    geom_violin(fill = "grey", color = NA, alpha = 0.5,
        adjust = 1.5, scale = "width", width = 0.3) +
    geom_quasirandom(width = 0.1, alpha = 0.5, size = 1) +
    scale_y_continuous("# of regulatory elements",
        limits = c(0, NA), labels = comma) +
    xlab("HPRC Fiber-seq pilot samples") +
    my_grid()
my_ggsave("Figures/union-total-res-per-sample.pdf", width = 1.6, height = 2)
```

# SD overlap analysis
```{r sd-analysis}
sd_df <- hap_wide[, .(
    n_total = .N,
    n_sd = sum(is_SD),
    frac_sd = sum(is_SD) / .N,
    total_hap_distinct = sum(haplotype_distinct != "Shared", na.rm = TRUE),
    n_sd_and_hap_distinct = sum(is_SD & haplotype_distinct != "Shared", na.rm = TRUE),
    frac_sd_and_hap_distinct = sum(is_SD & haplotype_distinct != "Shared", na.rm = TRUE) /
        max(sum(is_SD), 1),
    n_sd_and_genetic = sum(is_SD & haplotype_distinct == "Genetically haplotype-specific", na.rm = TRUE),
    frac_sd_and_genetic = sum(is_SD & haplotype_distinct == "Genetically haplotype-specific", na.rm = TRUE) /
        max(sum(is_SD), 1),
    n_sd_and_epigenetic = sum(is_SD & haplotype_distinct == "Epigenetically haplotype-selective", na.rm = TRUE),
    frac_sd_and_epigenetic = sum(is_SD & haplotype_distinct == "Epigenetically haplotype-selective", na.rm = TRUE) /
        max(sum(is_SD), 1)
), by = .(Individual_ID, ind_key)]
sd_df
```

# SD fraction beeswarm
```{r sd-frac-beeswarm}
sd_df %>%
    ggplot(aes(x = "", y = frac_sd)) +
    geom_violin(fill = "grey", color = NA, alpha = 0.5,
        adjust = 1.5, scale = "width", width = 0.3) +
    geom_quasirandom(color = "darkred",
        width = 0.1, alpha = 0.5, size = 1) +
    scale_y_continuous("REs that intersect SDs",
        limits = c(0, NA), labels = percent) +
    xlab("HPRC Fiber-seq pilot samples") +
    my_grid()
my_ggsave("Figures/union-sd-frac.pdf", width = 1.5, height = 2)
```

# SD REs by haplotype-distinct subcategory
```{r sd-hap-distinct-subcategory}
sd_names <- c(
    "Genetically haplotype-specific" = "frac_sd_and_genetic",
    "Epigenetically haplotype-selective" = "frac_sd_and_epigenetic",
    "Haplotype-distinct" = "frac_sd_and_hap_distinct"
)
sd_names_flip <- setNames(names(sd_names), sd_names)

sd_df %>%
    pivot_longer(
        cols = c(frac_sd_and_genetic, frac_sd_and_hap_distinct,
                 frac_sd_and_epigenetic),
        names_to = "type", values_to = "value"
    ) %>%
    mutate(
        type = sd_names_flip[type],
        type = factor(type, levels = rev(names(sd_names)))
    ) %>%
    ggplot(aes(x = value, y = type, color = type)) +
    geom_violin(fill = "grey", color = NA, alpha = 0.5,
        adjust = 1.5, scale = "width", width = 0.5) +
    geom_quasirandom(width = 0.25, alpha = 0.5, size = 1) +
    geom_text_repel(
        data = . %>%
            group_by(type) %>%
            summarise(median_frac = median(value, na.rm = TRUE)) %>%
            mutate(label = paste0(round(100 * median_frac, 1), "%")),
        aes(x = median_frac, label = label),
        min.segment.length = 0, nudge_y = 0.5,
        color = "black", size = 2
    ) +
    scale_y_discrete("") +
    scale_x_continuous("% of SD REs that are haplotype-distinct",
        limits = c(0, NA), labels = percent) +
    scale_color_manual(values = hdca_colors) +
    my_grid() +
    theme(legend.position = "none")
my_ggsave("Figures/union-sd-hap-distinct-frac.pdf", width = 3, height = 1.5)
```

# Density of H1-H2 diff by SD status
```{r diff-density-by-sd}
hap_wide[autosome == "Autosomes" & peak_status == "both" &
         pass_hap_coverage] %>%
    ggplot(aes(x = diff, fill = is_SD)) +
    geom_density(alpha = 0.5, adjust = 1) +
    scale_x_continuous("Difference in accessibility between haplotypes") +
    scale_y_continuous("Density") +
    scale_fill_manual("Intersects SD",
        values = c("TRUE" = "darkred", "FALSE" = "black")) +
    my_grid()
my_ggsave("Figures/union-diff-density-by-sd.pdf", width = 3, height = 2)
```

# ECDF of accessibility by SD status
```{r ecdf-by-sd}
set.seed(42)
hap_wide[, .SD[sample(.N, min(.N, 50000))], by = is_SD] %>%
    mutate(acc = (fire_coverage_1 + fire_coverage_2) /
                 (coverage_1 + coverage_2)) %>%
    ggplot(aes(x = acc, color = is_SD)) +
    stat_ecdf(linewidth = 0.75, alpha = 0.75) +
    scale_x_continuous("% actuation of regulatory elements",
        labels = percent) +
    scale_y_continuous("Cumulative % of regulatory elements",
        labels = percent) +
    scale_color_manual("Intersects SD",
        values = c("TRUE" = "darkred", "FALSE" = "black")) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-ecdf-by-sd.pdf", width = 2, height = 2)
```

# Density of accessibility by SD status
```{r accessibility-density-by-sd}
set.seed(42)
hap_wide[, .SD[sample(.N, min(.N, 50000))], by = is_SD] %>%
    mutate(acc = (fire_coverage_1 + fire_coverage_2) /
                 (coverage_1 + coverage_2)) %>%
    ggplot(aes(x = acc, fill = is_SD)) +
    geom_density(alpha = 0.5, adjust = 0.5) +
    scale_x_continuous("% actuation of regulatory elements",
        labels = percent) +
    scale_y_continuous("Density") +
    scale_fill_manual("Intersects SD",
        values = c("TRUE" = "darkred", "FALSE" = "black")) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-accessibility-density-by-sd.pdf",
    width = 2, height = 2)
```

# Size distribution by SD status
```{r size-by-sd}
set.seed(42)
hap_wide[peak_status == "both",
    .SD[sample(.N, min(.N, 50000))], by = is_SD
] %>%
    ggplot(aes(x = cons_end - cons_start, fill = is_SD)) +
    geom_density(alpha = 0.5, adjust = 0.5) +
    scale_x_continuous("Size of regulatory elements (bp)",
        labels = comma, trans = "log10", limits = c(100, 1000)) +
    annotation_logticks(sides = "b") +
    scale_y_continuous("Density") +
    scale_fill_manual("Intersects SD",
        values = c("TRUE" = "darkred", "FALSE" = "black")) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-size-by-sd.pdf", width = 2, height = 2)
```

# Fraction of haplotype-distinct REs in SDs
```{r hap-distinct-in-sds}
sd_df %>%
    ggplot(aes(y = Individual_ID,
               x = n_sd_and_hap_distinct / total_hap_distinct)) +
    geom_bar(stat = "identity", fill = "darkred", alpha = 0.7) +
    scale_x_continuous("Fraction of haplotype-distinct REs in SDs",
        limits = c(0, 1), labels = percent) +
    my_grid()
my_ggsave("Figures/union-frac-hap-distinct-in-sds.pdf",
    width = 3, height = 2)
```
