# HPRC2 ancestry PCA with Fiber-seq samples highlighted

Replicates the PCA space of HPRC2 Fig. 1d (bioRxiv 10.64898/2026.07.21.739710) and
marks the HPRC2 samples for which we have Fiber-seq data.

The space is PCLAI's reference PCA (Geleta et al. 2026, bioRxiv 10.64898/2026.03.23.713813):
1,561 low-admixture 1000G samples, two haplotypes each, embedded in PC1/PC2. HPRC2
haplotypes are not points in that reference, so a haplotype's position here is the
genome-wide mean of PCLAI's per-window predictions (`scripts/pclai-hap-means.sh`).
Haplotypes are plotted separately, so an admixed individual shows up as two points.
That averaging tracks an independent genotype PCA closely (Procrustes R^2 = 0.97 against
gnomAD, and no worse for admixed samples), but for recently admixed individuals the mean
is a centroid between ancestry clusters rather than a place any window sits;
`sd_x1`/`sd_x2` in the means table carry the per-window spread.

```{r setup}
source("Rscripts/utils.R")
library(mrvplot) # `pixi run setup` if this fails
```

```{r load-tables}
# 1000G reference cloud: 1,561 samples x 2 haplotypes in the PCLAI PC1/PC2 space.
ref_df <- fread("metadata/pclai_reference_pca.tsv")
setnames(ref_df, c("Sample", "Haplotype"), c("sample_id", "haplotype"))

# 1000G population -> superpopulation. Not used for colour, but it makes the
# table that mrv_ggsave() writes alongside the figure self-describing.
super_pop_df <- unique(
    fread("metadata/1kg_3202_samples_population.txt")[, .(
        sample_id = SampleID, pop = Population, super_pop = Superpopulation
    )]
)
ref_df <- merge(ref_df, super_pop_df, by = "sample_id", all.x = TRUE)

# Genome-wide mean PCLAI coordinate per HPRC2 haplotype (463 haplotypes).
hprc_df <- fread("metadata/pclai_hap_means.tsv")

# HPRC2 release 2 sample table.
hprc_meta <- fread("metadata/hprc2.metadata.tbl")[, .(
    sample_id = `Sample ID`,
    pop = `Population Abbreviation`,
    pop_descriptor = `Population Descriptor`
)]

# Which HPRC2 individuals we have Fiber-seq for, and on which platform.
fiberseq_df <- fread("metadata/fiberseq.metadata.tbl")[, .(
    platform = fcase(
        uniqueN(Platform) > 1, "PacBio + ONT",
        all(Platform == "PacBio"), "PacBio",
        default = "ONT"
    )
), by = .(sample_id = Individual_ID)]

hprc_df <- merge(hprc_df, hprc_meta, by = "sample_id", all.x = TRUE)
hprc_df <- merge(hprc_df, fiberseq_df, by = "sample_id", all.x = TRUE)

# GM12878 is in the Fiber-seq manifest but is not an HPRC2 sample, and HG00272 /
# GRCh38 are in HPRC2 but have no PCLAI predictions. Report rather than hide.
setdiff(fiberseq_df$sample_id, hprc_df$sample_id)
hprc_df[, .N, by = platform]
```

```{r hprc2-pca}
CLASS_COLORS <- c(
    "1000G reference" = "grey85",
    "HPRC2" = "grey45",
    "CHM13" = "forestgreen",
    "Fiber-seq" = "purple"
)
CLASS_ORDER <- names(CLASS_COLORS)

set_class <- function(dt, cls) dt[, class := factor(cls, levels = CLASS_ORDER)][]
ref_df <- set_class(ref_df, "1000G reference")
# CHM13 is a complete hydatidiform mole, so PCLAI releases one haplotype for it
# rather than two. Splitting it out keeps the HPRC2 count an even 2 per sample.
hprc_df[, class := factor(
    fcase(
        sample_id == "CHM13", "CHM13",
        !is.na(platform), "Fiber-seq",
        default = "HPRC2"
    ),
    levels = CLASS_ORDER
)]
chm13_df <- hprc_df[class == "CHM13"]
fs_df <- hprc_df[class == "Fiber-seq"]

# h1-h2 of the same individual, so admixed Fiber-seq samples read as a pair.
# Only drawn if every Fiber-seq sample has both haplotypes: a partial set of
# segments would read as a real difference between samples rather than as
# missing data. CHM13 is the one HPRC2 sample without a pair (it is haploid).
fs_pairs <- dcast(fs_df, sample_id + platform ~ haplotype, value.var = c("x1", "x2"))
hap_pairs_complete <- fs_df[, .N, by = sample_id][, all(N == 2)]
hap_pairs_complete

segment_layer <- if (hap_pairs_complete) {
    geom_segment(
        data = fs_pairs,
        aes(x = x1_1, y = x2_1, xend = x1_2, yend = x2_2),
        color = "black", linewidth = 0.2, alpha = 0.5,
        inherit.aes = FALSE
    )
}

# Every diploid HPRC2 haplotype is drawn grey; the Fiber-seq ones are then ringed
# in purple so the underlying point still shows through.
hprc_all <- copy(hprc_df[class != "CHM13"])[
    , class := factor("HPRC2", levels = CLASS_ORDER)
]

# Counts are haplotypes. n_hprc is the whole release, so the Fiber-seq count is a
# subset of it rather than a separate group. 463 not 464 because CHM13 is haploid.
n_ref <- nrow(ref_df)
n_hprc <- nrow(hprc_all)
n_fs <- nrow(fs_df)

# One table behind the figure so mrv_ggsave() journals it.
pca_df <- rbind(
    ref_df[, .(sample_id, haplotype, x1, x2, pop, super_pop, platform = NA_character_, class)],
    hprc_df[, .(sample_id, haplotype, x1, x2, pop, super_pop = NA_character_, platform, class)]
)
pca_df

p <- ggplot(pca_df, aes(x = x1, y = x2, color = class)) +
    geom_point(data = ref_df, size = 0.5, alpha = 0.5, shape = 16) +
    geom_point(data = hprc_all, size = 1, alpha = 0.7, shape = 16) +
    geom_point(data = chm13_df, size = 1, shape = 16) +
    # segment_layer +
    geom_point(
        data = fs_df,
        size = 1, shape = 21, fill = NA, stroke = 0.5
    ) +
    scale_color_manual(
        "",
        values = CLASS_COLORS, drop = FALSE,
        labels = c(
            "1000G reference" = glue("1000G reference (n = {n_ref})"),
            "HPRC2" = glue("HPRC2 (n = {n_hprc})"),
            "CHM13" = "CHM13 (n = 1)",
            "Fiber-seq" = glue("Fiber-seq (n = {n_fs})")
        ),
        guide = guide_legend(
            nrow = 2,
            override.aes = list(
                shape = c(16, 16, 16, 21), size = c(2, 2, 2, 2.2),
                fill = NA, stroke = 0.6, alpha = 1
            )
        )
    ) +
    scale_x_continuous("Mean PCLAI coordinate (reference PC1)") +
    scale_y_continuous("Mean PCLAI coordinate (reference PC2)") +
    coord_fixed() +
    theme_bw(base_size = 8) +
    theme(
        legend.position = "top",
        legend.key.size = unit(0.4, "cm"),
        panel.grid = element_blank(),
        aspect.ratio = 1
    )

mrv_ggsave("Figures/hprc2-pca.pdf", plot = p, width = 3, height = 3)
```
