# HSCA driver variants are enriched in the HLA region

Genome-wide view of where phylo-FIRE attributes haplotype-selective chromatin
accessibility (HSCA) to sequence. Forward-selection joint regression assigns driver
variants per tested peak; a peak is a *driver peak* if at least one variant is
attributed. The ideogram shows driver peaks piling up in the MHC beyond the
genome-wide rate, which is what motivates the HLA zoom-in.

Coordinates are CHM13 v2.0 (the consensus/graph reference the peaks are called in),
and the ideogram uses the CHM13 cytoband stains.

Tables come from `scripts/hsca-driver-peaks.sh` (phylo-FIRE run `hsca_hapsel` over
all HPRC Fiber-seq samples, plus the full consensus peak set from `mrv-figures`).

```{r setup}
source("Rscripts/utils.R")
library(mrvplot) # `pixi run setup` if this fails
library(karyoploteR)
```

```{r load-tables}
CHROMS <- paste0("chr", 1:22) # autosomes only

# Every consensus FIRE peak: the background, and the honest denominator for
# "is this region unusually sequence-driven".
all_peaks <- fread("metadata/all_peaks_chm13.tsv.gz")[chrom %in% CHROMS]

# The haplotype-selective peaks phylo-FIRE actually tested, with driver calls.
tested <- fread("metadata/hsca_driver_peaks.tsv.gz")[chrom %in% CHROMS]
drivers <- tested[has_driver == TRUE]

cat(
    "all peaks:", nrow(all_peaks),
    "| HSCA tested:", nrow(tested),
    "| driver peaks:", nrow(drivers), "\n"
)

# CHM13 ideogram scaffolding.
chrom_sizes <- fread("metadata/chm13v2.0.chrom.sizes",
    col.names = c("chrom", "size")
)[chrom %in% CHROMS]
chrom_sizes[, chrom := factor(chrom, levels = CHROMS)]
setorder(chrom_sizes, chrom)
chm13_genome <- toGRanges(data.frame(
    chr = as.character(chrom_sizes$chrom), start = 1, end = chrom_sizes$size
))
chm13_cyto <- toGRanges(as.data.frame(
    fread("metadata/chm13v2.0_cytobands.bed",
        col.names = c("chr", "start", "end", "name", "gieStain")
    )[chr %in% CHROMS]
))

# MHC in CHM13 coordinates, defined as GABBR1 through KIFC1 (3.80 Mb), matching
# how the driver-element counts are tabulated elsewhere in the project. Gene
# bounds from the UCSC hs1 catLiftOffGenesV1 track:
#   GABBR1 chr6:29,430,312-29,508,680   KIFC1 chr6:33,213,184-33,231,258
MHC <- list(chrom = "chr6", start = 29430312, end = 33231258)
```

```{r hla-enrichment}
# Two framings, because they answer different questions.
#   vs tested : given a haplotype-selective peak that was tested, is it more often
#               sequence-driven in the MHC? This is the driver-attribution claim.
#   vs all    : relative to all regulatory elements, does the MHC carry more
#               sequence-driven HSCA? This folds in the MHC being peak-dense.
in_mhc <- function(dt) dt$chrom == MHC$chrom & dt$start >= MHC$start & dt$end <= MHC$end

fisher_row <- function(k_in, n_in, k_out, n_out, label) {
    tab <- matrix(c(k_in, n_in - k_in, k_out, n_out - k_out),
        nrow = 2,
        dimnames = list(c("driver", "not"), c("MHC", "rest"))
    )
    ft <- fisher.test(tab)
    data.table(
        comparison = label,
        mhc = sprintf("%d/%d (%.1f%%)", k_in, n_in, 100 * k_in / n_in),
        rest = sprintf("%d/%d (%.1f%%)", k_out, n_out, 100 * k_out / n_out),
        odds_ratio = round(unname(ft$estimate), 2),
        ci = sprintf("%.2f-%.2f", ft$conf.int[1], ft$conf.int[2]),
        p = signif(ft$p.value, 3)
    )
}

tested[, mhc := in_mhc(tested)]
all_peaks[, mhc := in_mhc(all_peaks)]

enrichment <- rbind(
    fisher_row(
        tested[mhc == TRUE, sum(has_driver)], tested[mhc == TRUE, .N],
        tested[mhc == FALSE, sum(has_driver)], tested[mhc == FALSE, .N],
        "drivers / HSCA peaks tested"
    ),
    fisher_row(
        tested[mhc == TRUE, sum(has_driver)], all_peaks[mhc == TRUE, .N],
        tested[mhc == FALSE, sum(has_driver)], all_peaks[mhc == FALSE, .N],
        "drivers / all FIRE peaks"
    ),
    fisher_row(
        tested[mhc == TRUE, .N], all_peaks[mhc == TRUE, .N],
        tested[mhc == FALSE, .N], all_peaks[mhc == FALSE, .N],
        "HSCA tested / all FIRE peaks"
    )
)
enrichment
```

```{r rolling-window-test}
# Scan for regions where tested peaks are unusually often sequence-driven.
#
# The window is a fixed NUMBER OF PEAKS, not a fixed number of base pairs. Peak
# density varies by more than an order of magnitude along the genome, so fixed-bp
# windows carry wildly unequal power: a 5 Mb window held anywhere from 20 to 200+
# tested peaks, and the binomial test was correspondingly noisy in sparse ones.
# A fixed peak count gives every window the same n and makes the p-values
# comparable across the genome.
ROLL_PEAKS <- 200 # tested peaks per window
ROLL_STEP <- 50 # peaks between window starts

GENOME_RATE <- nrow(drivers) / nrow(tested)

setorder(tested, chrom, start)
roll <- tested[,
    {
        n <- .N
        if (n < ROLL_PEAKS) {
            NULL
        } else {
            # frollsum is right-aligned: element i holds the sum over (i-k+1):i.
            hits <- frollsum(as.numeric(has_driver), ROLL_PEAKS)
            i <- seq(ROLL_PEAKS, n, by = ROLL_STEP)
            .(
                win_start = start[i - ROLL_PEAKS + 1], win_end = end[i],
                mid = (start[i - ROLL_PEAKS + 1] + end[i]) / 2,
                n_drivers = hits[i]
            )
        }
    },
    by = chrom
]

roll[, `:=`(
    n_tested = ROLL_PEAKS,
    rate = n_drivers / ROLL_PEAKS,
    span_mb = (win_end - win_start) / 1e6
)]
# One-sided binomial upper tail, vectorised; equivalent to binom.test(alternative
# = "greater") but cheap enough to run over every window.
roll[, p := pbinom(n_drivers - 1, n_tested, GENOME_RATE, lower.tail = FALSE)]
roll[, fdr := p.adjust(p, method = "BH")]

# Windows are overlapping, so BH here is a screen for where to look, not a clean
# family-wise statement. The MHC claim rests on the pre-specified test below.
roll[, overlaps_mhc := chrom == MHC$chrom & win_start <= MHC$end & win_end >= MHC$start]

cat(
    "rolling windows:", nrow(roll), "of", ROLL_PEAKS, "peaks each;",
    "median span", round(median(roll$span_mb), 1), "Mb\n"
)
cat("windows at FDR < 0.05:", roll[fdr < 0.05, .N], "\n\n")

cat("top windows by driver rate:\n")
print(head(roll[order(-rate), .(chrom,
    Mb = round(mid / 1e6, 1), span_mb = round(span_mb, 1),
    n_drivers, rate = round(rate, 3), fdr = signif(fdr, 2), overlaps_mhc
)], 10))

cat("\nwindows overlapping the MHC:\n")
print(roll[overlaps_mhc == TRUE, .(chrom,
    Mb = round(mid / 1e6, 1), span_mb = round(span_mb, 1),
    n_drivers, rate = round(rate, 3), fdr = signif(fdr, 2)
)])
```


```{r normalised-density}
# Densities on a common grid, each scaled to sum to 1 so "all peaks" and "driver
# peaks" enclose the same total area. That turns the comparison into "where are
# drivers concentrated RELATIVE to where peaks are", instead of the raw counts,
# where all peaks (566k) would dwarf drivers (3.2k).
#
# Sliding, not tiled: counts are binned at DENSITY_STEP and then summed over
# DENSITY_WINDOW / DENSITY_STEP consecutive bins, so each plotted point is a
# 5 Mb window and consecutive points overlap by 4.5 Mb. Overlap inflates every
# curve by the same factor, so the equal-area comparison between curves is
# unaffected. The sum is right-aligned with the centre computed explicitly,
# which stays correct for an even DENSITY_K where align = "center" would not.
#
# This grid is for PLOTTING only. The statistical scan uses rolling windows of a
# fixed peak count instead, since peak density varies far too much for a fixed-bp
# window to carry constant power.
DENSITY_WINDOW <- 5e6
DENSITY_STEP <- 5e5
DENSITY_K <- DENSITY_WINDOW / DENSITY_STEP
stopifnot(DENSITY_WINDOW %% DENSITY_STEP == 0)

grid <- chrom_sizes[, .(window = seq(0, size, by = DENSITY_STEP)), by = chrom]

bin_density <- function(dt, label) {
    b <- dt[, .N, by = .(chrom, window = floor(start / DENSITY_STEP) * DENSITY_STEP)]
    d <- merge(grid, b, by = c("chrom", "window"), all.x = TRUE)
    d[is.na(N), N := 0L]
    setorder(d, chrom, window)
    # Windowed count via cumulative sums rather than frollsum: frollsum has to
    # fill the leading DENSITY_K - 1 bins, and filling them with 0 makes the
    # first 4.5 Mb of every chromosome read as empty. cumsum gives those bins
    # their true partial-window count instead.
    d[, cs := cumsum(N), by = chrom]
    # data.table::shift qualified: utils.R loads IRanges, which masks shift().
    d[, n_win := cs - data.table::shift(cs, DENSITY_K, fill = 0), by = chrom]
    # Window covers [w + STEP - WINDOW, w + STEP), clamped to the chromosome
    # start so the centre of a partial window is not reported off the sequence.
    d[, win_start := pmax(0, window + DENSITY_STEP - DENSITY_WINDOW)]
    d[, `:=`(
        frac = n_win / sum(n_win),
        mid = (win_start + window + DENSITY_STEP) / 2,
        track = label
    )]
    d[, cs := NULL]
    d[]
}

dens_all <- bin_density(all_peaks, "all peaks")
dens_tested <- bin_density(tested, "HSCA tested")
dens_drivers <- bin_density(drivers, "driver peaks")
dens <- rbind(dens_all, dens_tested, dens_drivers)
dens[, .(total_area = sum(frac), max_frac = round(max(frac), 4)), by = track]
```

```{r karyotype-helper}
# Shared CHM13 karyotype for both figures.
#   chromosomes = "all": a custom genome has no canonical-chromosome list, so
#     without it karyoploteR heuristically filters and warns. chm13_genome is
#     already restricted to the autosomes.
#   cyto_colors["border"]: at whole-genome scale each band is a couple of pixels
#     wide, so the default black band borders swamp the stain fill and the
#     ideogram renders solid black. The entry must be overridden in place --
#     appending a second "border" with c() does nothing, because
#     color.table["border"] matches the built-in one first.
base_karyotype <- function(pp, label = "MHC") {
    cyto_colors <- getCytobandColors(color.schema = "circos")
    cyto_colors["border"] <- "transparent"
    kp <- plotKaryotype(
        genome = chm13_genome, cytobands = chm13_cyto, chromosomes = "all",
        plot.type = 4, cex = 0.3, plot.params = pp, ideogram.plotter = NULL
    )
    kpAddCytobands(kp, color.table = cyto_colors)
    # Mark the MHC with a tick at the base of the data panel plus a label, rather
    # than a band across it. At genome scale 4.8 Mb is about two pixels wide, so a
    # band saturated enough to read also covers the chr6 spike it points at, and
    # widening it to compensate would misrepresent the interval.
    kpRect(kp,
        chr = MHC$chrom, x0 = MHC$start, x1 = MHC$end, y0 = 0, y1 = 1,
        col = "#e69138", border = "#e69138", lwd = 1.5, r0 = -0.02, r1 = 0.06
    )
    # Label above the data; inside the panel it gets buried under the density fill.
    kpText(kp,
        chr = MHC$chrom, x = mean(c(MHC$start, MHC$end)), y = 1.06,
        labels = label, col = "#e69138", cex = 0.75
    )
    kp
}


FIG <- "Figures/hsca-drivers-ideogram.pdf"

draw_ideogram <- function() {
    pp <- getDefaultPlotParams(plot.type = 4)
    pp$data1height <- 300 # data vs ideogram balance
    pp$ideogramheight <- 30 # default 10 is a hairline at this width
    pp$ideogramlateralmargin <- 0.0005 # gap between chromosomes; default 0.003
    pp$leftmargin <- 0.03
    pp$rightmargin <- 0.01
    pp$bottommargin <- 18
    pp$topmargin <- 42 # room for the MHC label above the data

    # Fisher result folded into the MHC label as one centred string, read from
    # the enrichment table so the figure cannot drift out of sync with the test.
    # Two kpText calls do not line up: kpText centres vertically on y, while
    # pos = 4 uses a baseline offset, so the two strings sit at different
    # heights and butt together horizontally with no gap.
    # The main figure plots drivers against all peaks, so that is the row quoted.
    e <- enrichment[comparison == "drivers / all FIRE peaks"]
    ex <- floor(log10(e$p))
    mant <- round(e$p / 10^ex, 1)
    kp <- base_karyotype(pp, label = bquote(
        bold("MHC") ~ ~ OR == .(sprintf("%.1f", e$odds_ratio)) *
            "," ~ italic(P) == .(mant) %*% 10^.(ex)
    ))

    # One data row, three area-normalised densities overlaid. Each curve encloses
    # the same total area, so height is "share of this class found here" and the
    # curves are directly comparable despite 566k vs 23k vs 3.2k peaks. Where red
    # clears grey, drivers are concentrated beyond where peaks simply are.
    # Scale to the two curves actually drawn; HSCA tested is a supplementary
    # comparison and stays out of the main figure.
    dmax <- max(dens_all$frac, dens_drivers$frac)
    area <- function(d, fill, line) {
        kpArea(kp,
            chr = d$chrom, x = d$mid, y = d$frac / dmax,
            r0 = 0.02, r1 = 1, col = fill, border = line, lwd = 0.3
        )
    }
    area(dens_all, "grey60", "grey35")
    area(dens_drivers, paste0(Red, "66"), Red)
    kpAddLabels(kp, "peak density", r0 = 0.02, r1 = 1, cex = 0.7, srt = 90, pos = 3)
    legend("topleft",
        legend = c("all peaks", "driver peaks"),
        fill = c("grey60", paste0(Red, "66")),
        border = c("grey35", Red), bty = "n", cex = 0.7, horiz = TRUE, xpd = NA
    )
}

# karyoploteR draws to a device rather than returning a ggplot, so mrv_ggsave()
# cannot be used. Do what it does by hand: write the figure, copy it to tmp.pdf
# for previewing over VS Code/SSH, and journal the data behind it.
save_karyoplot <- function(file, draw, width, height, tables = list(), tmp = TRUE) {
    dir.create(dirname(file), showWarnings = FALSE, recursive = TRUE)
    dir.create(file.path(dirname(file), "Tables"), showWarnings = FALSE, recursive = TRUE)
    pdf(file, width = width, height = height)
    draw()
    invisible(dev.off())
    # Only the main figure claims tmp.*, so the preview always shows the figure
    # being iterated on rather than whichever chunk happened to run last.
    if (tmp) file.copy(file, paste0("tmp.", tools::file_ext(file)), overwrite = TRUE)
    for (nm in names(tables)) {
        fwrite(tables[[nm]],
            file.path(dirname(file), "Tables", paste0(nm, ".tbl.gz")),
            sep = "\t"
        )
    }
    print(file)
}

save_karyoplot(FIG, draw_ideogram, width = 12, height = 2, tables = list(
    # The curves the figure actually draws: one row per class per window.
    "hsca-drivers-ideogram" = dens[, .(track, chrom, window, mid, n = n_win, frac)],
    # Statistics kept separate from the plotted data.
    "hsca-drivers-rolling-test" = roll,
    "hsca-drivers-enrichment" = enrichment
))
```

# Supplementary: driver rate under two denominators

The main figure normalises by area, which shows *where* drivers concentrate but not
*how often* a peak is sequence-driven. These two tracks divide the same driver counts
by the two denominators the question can take:

- **/ HSCA tested** — of the haplotype-selective peaks phylo-FIRE tested, what fraction
  got a driver attributed. Conditions on the peak already being haplotype-selective.
- **/ all FIRE peaks** — of every regulatory element, what fraction is a sequence-driven
  HSCA peak. Folds in the MHC being both peak-dense and unusually haplotype-selective.

```{r rate-denominators}
# Per-window rates on the same sliding grid, for the journalled table. n_win is
# the windowed count, so these are 2.5 Mb windows stepping 0.5 Mb.
rates <- Reduce(
    function(a, b) merge(a, b, by = c("chrom", "window")),
    list(
        dens_drivers[, .(chrom, window, mid, n_drivers = n_win)],
        dens_tested[, .(chrom, window, n_tested = n_win)],
        dens_all[, .(chrom, window, n_all = n_win)]
    )
)

# Windows with too few peaks in the denominator are dropped rather than reported
# as 0/1 noise, separately for each denominator.
MIN_TESTED_PER_WINDOW <- 10
MIN_ALL_PER_WINDOW <- 100
rates[, `:=`(
    rate_hsca = fifelse(n_tested >= MIN_TESTED_PER_WINDOW, n_drivers / n_tested, NA_real_),
    rate_all = fifelse(n_all >= MIN_ALL_PER_WINDOW, n_drivers / n_all, NA_real_)
)]

BASE_HSCA <- nrow(drivers) / nrow(tested)
BASE_ALL <- nrow(drivers) / nrow(all_peaks)
cat(sprintf(
    "genome-wide: %.1f%% of HSCA tested peaks, %.2f%% of all peaks\n",
    100 * BASE_HSCA, 100 * BASE_ALL
))
rates[
    chrom == MHC$chrom & window >= MHC$start - 2e6 & window <= MHC$end + 2e6,
    .(chrom, window, n_drivers, n_tested, n_all,
        rate_hsca = round(rate_hsca, 3), rate_all = round(rate_all, 4)
    )
]
```

```{r hsca-drivers-denominators}
SUPP <- "Figures/hsca-drivers-denominators.pdf"

draw_denominators <- function() {
    pp <- getDefaultPlotParams(plot.type = 4)
    pp$data1height <- 300 # data vs ideogram balance
    pp$ideogramheight <- 40 # default 10 is a hairline at this width
    pp$ideogramlateralmargin <- 0.0005 # gap between chromosomes; default 0.003
    pp$leftmargin <- 0.03
    pp$rightmargin <- 0.01
    pp$bottommargin <- 18
    pp$topmargin <- 42 # room for the MHC label above the data
    kp <- base_karyotype(pp)

    # Same overlay as the main figure, run against each denominator in turn.
    # Every curve is area-normalised, so within a row the two are directly
    # comparable and the question is just "does red clear the background".
    overlay <- function(bg, bg_fill, bg_line, label, r0, r1) {
        ymax <- max(bg$frac, dens_drivers$frac)
        kpArea(kp,
            chr = bg$chrom, x = bg$mid, y = bg$frac / ymax,
            r0 = r0, r1 = r1, col = bg_fill, border = bg_line, lwd = 0.3
        )
        kpArea(kp,
            chr = dens_drivers$chrom, x = dens_drivers$mid,
            y = dens_drivers$frac / ymax,
            r0 = r0, r1 = r1, col = paste0(Red, "66"), border = Red, lwd = 0.3
        )
        kpAddLabels(kp, label, r0 = r0, r1 = r1, cex = 0.7, srt = 90, pos = 3)
    }

    overlay(dens_tested, paste0(Teal, "66"), Teal, "vs HSCA tested", 0.56, 1)
    overlay(dens_all, "grey60", "grey35", "vs all peaks", 0.02, 0.46)
    legend("topleft",
        legend = c("HSCA tested", "all peaks", "driver peaks"),
        fill = c(paste0(Teal, "66"), "grey60", paste0(Red, "66")),
        border = c(Teal, "grey35", Red), bty = "n", cex = 0.7, horiz = TRUE, xpd = NA
    )
}

save_karyoplot(SUPP, draw_denominators, width = 7.5, height = 3.2, tmp = FALSE, tables = list(
    "hsca-drivers-denominators" = rates
))
```
