```{r setup}
source("Rscripts/utils.R")
# Column definitions: see Tables/union-peaks-README.md
```

# Load union peaks (consensus coords, all sites)
```{r load-data}
union_df <- my_read_bed(
    cmd = "bgzip -dc -@ 8 Tables/union-peaks-cons.bed.gz",
    sort = FALSE
)
cat("Total rows:", nrow(union_df), "\n")
```

# Restrict to individuals with both PacBio and ONT
```{r both-platforms}
plats_per_ind <- union_df[
    !is.na(Platform),
    .(platforms = paste(sort(unique(Platform)), collapse = ",")),
    by = Individual_ID
]
ind_with_both <- plats_per_ind[platforms == "ONT,PacBio", Individual_ID]
cat("Individuals with both PacBio and ONT:", length(ind_with_both), "\n")
print(ind_with_both)

df <- union_df[Individual_ID %in% ind_with_both & !is.na(Platform)]
cat("Rows after restricting to those individuals:", nrow(df), "\n")
```

# Aggregate over haplotypes per (Individual_ID, Platform, consensus_peak_id)
```{r aggregate}
# Sum coverage/fire across both haplotypes so the per-hap label (which may not be
# comparable across platforms) doesn't matter for a genome-wide comparison.
df_agg <- df[, .(
    coverage = sum(coverage, na.rm = TRUE),
    fire_coverage = sum(fire_coverage, na.rm = TRUE),
    is_peak = any(is_peak, na.rm = TRUE)
), by = .(Individual_ID, Platform, consensus_peak_id)]
df_agg[, fire_cov_OV_cov := fire_coverage / coverage]
cat("Aggregated rows:", nrow(df_agg), "\n")
```

# Pivot PacBio vs ONT and filter to peaks in at least one platform
```{r pivot-and-filter}
MIN_COV <- 30

key_cols <- c("Individual_ID", "consensus_peak_id")
val_cols <- c("coverage", "fire_coverage", "fire_cov_OV_cov", "is_peak")

pb <- df_agg[Platform == "PacBio", c(key_cols, val_cols), with = FALSE]
ont <- df_agg[Platform == "ONT", c(key_cols, val_cols), with = FALSE]
setnames(pb, val_cols, paste0(val_cols, "_pb"))
setnames(ont, val_cols, paste0(val_cols, "_ont"))

pb_ont <- merge(pb, ont, by = key_cols)
cat("Sites measured in both platforms:", nrow(pb_ont), "\n")

# is_peak in PacBio OR ONT for this individual
pb_ont <- pb_ont[is_peak_pb == TRUE | is_peak_ont == TRUE]
cat("Sites that are a peak in PacBio or ONT:", nrow(pb_ont), "\n")

# require >= MIN_COV total coverage in both platforms (summed across haps)
pb_ont <- pb_ont[coverage_pb >= MIN_COV & coverage_ont >= MIN_COV]
cat("Sites with >=", MIN_COV, "coverage in both platforms:", nrow(pb_ont), "\n")
```

# Per-individual Pearson correlations (genome-wide)
```{r correlations}
correlations <- pb_ont[,
    {
        r <- cor(fire_cov_OV_cov_pb, fire_cov_OV_cov_ont, method = "pearson")
        n <- .N
        # cor.test() p-values underflow to 0 at huge n; compute log p on the
        # t-distribution directly and report log10(p) so it stays informative.
        t_stat <- r * sqrt(n - 2) / sqrt(1 - r^2)
        log_p <- pt(-abs(t_stat), df = n - 2, log.p = TRUE) + log(2)
        .(
            n_sites = n,
            pearson_r = r,
            p_value = exp(log_p),
            log10_p = log_p / log(10)
        )
    },
    by = Individual_ID
][order(Individual_ID)]

correlations
fwrite(
    correlations,
    "Tables/pacbio-vs-ont-correlations.tsv",
    sep = "\t", quote = FALSE
)
cat("Saved Tables/pacbio-vs-ont-correlations.tsv\n")
```

# Scatter: accessibility (PacBio vs ONT) per individual
```{r scatter-accessibility}
pb_ont %>%
    ggplot(aes(x = fire_cov_OV_cov_pb, y = fire_cov_OV_cov_ont)) +
    geom_hex(bins = 40) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    stat_cor(method = "pearson", size = 2) +
    facet_wrap(~Individual_ID) +
    scale_x_continuous("PacBio accessibility (fire/coverage)",
        labels = percent, limits = c(0, 1)
    ) +
    scale_y_continuous("ONT accessibility (fire/coverage)",
        labels = percent, limits = c(0, 1)
    ) +
    coord_equal() +
    my_grid()
my_ggsave(
    "Figures/pacbio-vs-ont-accessibility-scatter.pdf",
    width = 6, height = 6
)
```

# Histogram: PacBio - ONT accessibility delta per individual
```{r delta-histogram}
delta_summary <- pb_ont[, .(
    median_delta = median(fire_cov_OV_cov_pb - fire_cov_OV_cov_ont,
        na.rm = TRUE
    ),
    mean_delta = mean(fire_cov_OV_cov_pb - fire_cov_OV_cov_ont,
        na.rm = TRUE
    )
), by = Individual_ID]
delta_summary

pb_ont %>%
    mutate(delta = fire_cov_OV_cov_pb - fire_cov_OV_cov_ont) %>%
    ggplot(aes(x = delta)) +
    geom_histogram(
        binwidth = 0.02, fill = "grey70", color = "black",
        linewidth = 0.2
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    geom_vline(
        data = delta_summary,
        aes(xintercept = mean_delta),
        color = "blue", linetype = "dotted"
    ) +
    geom_text(
        data = delta_summary,
        aes(
            x = mean_delta,
            y = Inf,
            label = paste0("mean = ", percent(mean_delta, accuracy = 0.1))
        ),
        color = "blue", size = 2, hjust = -0.05, vjust = 1.5
    ) +
    facet_wrap(~Individual_ID, scales = "free_y") +
    scale_x_continuous(
        "PacBio - ONT accessibility",
        labels = percent, limits = c(-1, 1)
    ) +
    scale_y_continuous("# peaks", labels = comma) +
    my_grid()
my_ggsave(
    "Figures/pacbio-vs-ont-accessibility-delta-hist.pdf",
    width = 6, height = 4
)
```


```{r}
pb_ont %>%
    group_by(Individual_ID) %>%
    summarise(
        mean(coverage_pb),
        median(coverage_pb),
        mean(coverage_ont),
        median(coverage_ont)
    )
```