```{r setup}
source("Rscripts/utils.R")
```

# Load paralog pairs
```{r load-data}
pairs <- fread(cmd = "bgzip -cd -@ 16 paralog_pairs_enriched.tsv.gz")
cat("Paralog pairs:", nrow(pairs), "\n")
cat("Unique paralog groups:", uniqueN(pairs$paralog_group_id_query), "\n")
cat("Unique samples:", uniqueN(pairs$paralog1_sample_id), "\n")
glimpse(pairs)
# table(pairs$paralog1_sample_id, pairs$paralog2_sample_id)
```


# Get allelic pairs: extract each paralog as its own row, keeping both hap stats
```{r allelic-pairs}
extract_paralog <- function(dt, paralog = "paralog1") {
    p <- paralog
    dt[, .(
        consensus_peak_id = get(paste0(p, "_consensus_peak_id")),
        sample_id = get(paste0(p, "_sample_id")),
        haplotype_sample = get(paste0(p, "_haplotype_sample")),
        fire_coverage_1 = get(paste0(p, "_fire_coverage_1")),
        fire_coverage_2 = get(paste0(p, "_fire_coverage_2")),
        coverage_1 = get(paste0(p, "_coverage_1")),
        coverage_2 = get(paste0(p, "_coverage_2")),
        acc_1 = fifelse(
            get(paste0(p, "_coverage_1")) > 0,
            get(paste0(p, "_fire_coverage_1")) / get(paste0(p, "_coverage_1")),
            NA_real_
        ),
        acc_2 = fifelse(
            get(paste0(p, "_coverage_2")) > 0,
            get(paste0(p, "_fire_coverage_2")) / get(paste0(p, "_coverage_2")),
            NA_real_
        ),
        fire_ratio = get(paste0(p, "_fire_ratio")),
        is_peak = get(paste0(p, "_is_peak")),
        is_SD = get(paste0(p, "_is_SD")),
        haplotype_distinct = get(paste0(p, "_haplotype_distinct")),
        paralog = paralog,
        paralog_group = paralog_group_id_query,
        lifted_blast_identity,
        delta_fire_ratio,
        qc_clean
    )]
}

allelic_pairs <- rbind(
    extract_paralog(pairs, "paralog1"),
    extract_paralog(pairs, "paralog2")
)

cat("Allelic rows:", nrow(allelic_pairs), "\n")
cat("(should be ~2x original:", nrow(pairs), "* 2 =", nrow(pairs) * 2, ")\n")
allelic_pairs <- allelic_pairs %>%
    filter(
        !is.na(fire_coverage_1),
        !is.na(fire_coverage_2),
        !is.na(coverage_1),
        !is.na(coverage_2)
    )
```

# QC: coverage and peak stats
```{r qc-stats}
cat("Both are peaks:", sum(pairs$paralog1_is_peak & pairs$paralog2_is_peak, na.rm = TRUE), "\n")
cat("One is peak:", sum(xor(pairs$paralog1_is_peak, pairs$paralog2_is_peak), na.rm = TRUE), "\n")
cat("Neither is peak:", sum(!pairs$paralog1_is_peak & !pairs$paralog2_is_peak, na.rm = TRUE), "\n")

cat("\nBoth in SD:", sum(pairs$paralog1_is_SD & pairs$paralog2_is_SD, na.rm = TRUE), "\n")
cat("QC clean:", sum(pairs$qc_clean, na.rm = TRUE), "\n")
```

# Filter to QC-clean pairs
```{r filter}
clean_pairs <- pairs[qc_clean == TRUE]
cat("QC-clean pairs:", nrow(clean_pairs), "\n")
```

# Sequence identity between paralogs
```{r seq-identity}
clean_pairs %>%
    ggplot(aes(x = lifted_blast_identity)) +
    stat_bin(geom = "step", binwidth = 0.005, position = "identity") +
    scale_x_continuous("BLAST identity between paralog pair",
        labels = percent, limits = c(0.8, 1)
    ) +
    scale_y_continuous("# of paralog pairs", labels = comma) +
    my_grid()
my_ggsave("Figures/paralog-pairs-identity.pdf", width = 3, height = 2)
```

# Delta fire ratio: accessibility difference between paralogs
```{r delta-fire-ratio}
# take all the alleles for all the cons IDs in a paralog group
# then look at the distribution of FIRE scores
# compare that to the distribution of paralog FIRE scores
# can we use sequence to explain why the paralogs are more differentially accesible than the alleles?
mean(allelic_pairs$coverage_1)
mean(clean_pairs$paralog1_coverage, na.rm = TRUE)
allelic_pairs %>%
    mutate(
        type = "allelic",
        delta_fire_ratio = fire_coverage_1 / coverage_1 - fire_coverage_2 / coverage_2
    ) %>%
    select(type, delta_fire_ratio) %>%
    bind_rows(
        clean_pairs %>%
            mutate(
                delta_fire_ratio = paralog1_fire_coverage / paralog1_coverage - paralog2_fire_coverage / paralog2_coverage,
                # testable = paralog1_fire_coverage + paralog2_fire_coverage > 5,
                type = "paralog"
            ) %>%
            filter(
                # testable,
                !is.na(paralog1_coverage),
                !is.na(paralog2_coverage),
            ) %>%
            select(type, delta_fire_ratio)
    ) %>%
    ggplot(aes(x = delta_fire_ratio, color = type)) +
    stat_bin(
        aes(y = after_stat(density)),
        geom = "step",
        binwidth = 0.02, position = "identity"
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_x_continuous("Accessibility difference between paralogs",
        limits = c(-1, 1)
    ) +
    scale_y_continuous("# of paralog pairs", labels = comma) +
    my_grid()
my_ggsave("Figures/paralog-pairs-delta-fire.pdf", width = 3, height = 2)
```

# Accessibility of paralog1 vs paralog2
```{r paralog-scatter}
clean_pairs %>%
    ggplot(aes(x = paralog1_fire_ratio, y = paralog2_fire_ratio)) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous("Paralog 1 accessibility", labels = percent) +
    scale_y_continuous("Paralog 2 accessibility", labels = percent) +
    coord_equal() +
    my_grid()
my_ggsave("Figures/paralog-pairs-scatter.pdf", width = 3, height = 3)
```

# Identity vs accessibility difference
```{r identity-vs-delta}
clean_pairs %>%
    ggplot(aes(x = lifted_blast_identity, y = abs(delta_fire_ratio))) +
    geom_hex(bins = 50) +
    scale_fill_distiller("", palette = "Spectral", trans = "log10") +
    scale_x_continuous("Sequence identity between paralogs", labels = percent) +
    scale_y_continuous("| Accessibility difference |", labels = percent) +
    my_grid()
my_ggsave("Figures/paralog-pairs-identity-vs-delta.pdf", width = 3, height = 2)
```

# Pairs where one is a peak and the other is not
```{r discordant-peaks}
clean_pairs[, discordant := xor(
    paralog1_is_peak == TRUE,
    paralog2_is_peak == TRUE
)]

cat("Discordant pairs (one peak, one not):", sum(clean_pairs$discordant), "\n")
cat("Fraction discordant:", round(mean(clean_pairs$discordant) * 100, 1), "%\n")

clean_pairs %>%
    ggplot(aes(x = discordant, y = abs(delta_fire_ratio))) +
    geom_violin(fill = "grey90") +
    geom_quasirandom(method = "pseudorandom", size = 0.2, alpha = 0.1) +
    scale_y_continuous("| Accessibility difference |", labels = percent) +
    scale_x_discrete("", labels = c("TRUE" = "Discordant", "FALSE" = "Concordant")) +
    my_grid()
my_ggsave("Figures/paralog-pairs-discordant.pdf", width = 2, height = 2)
```

# Per paralog group summary
```{r per-group}
group_summary <- clean_pairs[, .(
    n_pairs = .N,
    n_samples = uniqueN(paralog1_haplotype_sample),
    mean_identity = mean(lifted_blast_identity, na.rm = TRUE),
    mean_delta = mean(abs(delta_fire_ratio), na.rm = TRUE),
    frac_discordant = mean(discordant)
), by = paralog_group_id_query]

group_summary %>%
    ggplot(aes(x = mean_identity, y = mean_delta)) +
    geom_point(aes(size = n_samples), alpha = 0.5) +
    scale_x_continuous("Mean sequence identity", labels = percent) +
    scale_y_continuous("Mean | accessibility difference |", labels = percent) +
    scale_size_continuous("# samples") +
    my_grid() +
    theme(legend.position = "right")
my_ggsave("Figures/paralog-pairs-group-summary.pdf", width = 3, height = 2.5)
```
