```{r setup}
source("Rscripts/utils.R")
```

# Load data
```{r load-data}
revived_genes <- fread("../pseduo-genes/pseudogene_with_strand.tsv",
    col.names = c("sample_hap", "contig", "start", "end", "gene_name", "strand")
)
dead_genes <- fread("../pseduo-genes/real_pseudogene_with_strand.tsv",
    col.names = c("sample_hap", "contig", "start", "end", "gene_name", "strand")
)
positive_control_genes <- fread("../pseduo-genes/protein_coding_strand.txt.gz",
    col.names = c("sample_hap", "contig", "start", "end", "strand")
)
# protein-coding file has no gene_name column; add placeholder so the
# downstream pipeline (which carries gene_name through) keeps working.
positive_control_genes[, gene_name := NA_character_]

peaks <- my_read_bed(cmd = "bgzip -dc -@ 8 Tables/peaks-asm.bed.gz", sort = FALSE)
all_sites <- my_read_bed(cmd = "bgzip -dc -@ 8 Tables/union-peaks-asm.bed.gz", sort = FALSE)

# parse sample_hap and build asm_chr-compatible contig name
parse_genes <- function(dt) {
    dt[, `:=`(
        Individual_ID = sub("_h[12]$", "", sample_hap),
        Haplotype = sub(".*_h", "", sample_hap),
        # match peaks asm_chr format: ID#hap#contig
        asm_chr = paste0(
            sub("_h[12]$", "", sample_hap), "#",
            sub(".*_h", "", sample_hap), "#",
            contig
        )
    )]
}
parse_genes(revived_genes)
parse_genes(dead_genes)
parse_genes(positive_control_genes)

cat("Revived genes:", nrow(revived_genes), "across", uniqueN(revived_genes$gene_name), "genes\n")
cat("Dead genes:", nrow(dead_genes), "across", uniqueN(dead_genes$gene_name), "genes\n")
cat("Positive control (protein-coding) loci:", nrow(positive_control_genes), "\n")
```

# Filter to samples with fiber-seq data
```{r filter-samples}
fiber_samples <- peaks[, unique(Individual_ID)]
cat("Fiber-seq individuals:", length(fiber_samples), "\n")

revived_fiber <- revived_genes[Individual_ID %in% fiber_samples]
dead_fiber <- dead_genes[Individual_ID %in% fiber_samples]
positive_fiber <- positive_control_genes[Individual_ID %in% fiber_samples]

length(unique(revived_fiber$Individual_ID))

cat(
    "Revived genes in fiber-seq samples:", nrow(revived_fiber),
    "across", uniqueN(revived_fiber$gene_name), "genes\n"
)
cat(
    "Dead genes in fiber-seq samples:", nrow(dead_fiber),
    "across", uniqueN(dead_fiber$gene_name), "genes\n"
)
cat(
    "Positive control loci in fiber-seq samples:", nrow(positive_fiber), "\n"
)
```

# Intersect gene promoters with peaks
```{r intersect-peaks}
# TSS +/- 500bp based on strand
FLANK <- 500

make_regions <- function(dt) {
    dt[, .(
        chrom = asm_chr,
        full_start = pmax(start - FLANK, 0L),
        full_end = end + FLANK,
        start = pmax(fifelse(strand == "+", start - FLANK, end - FLANK), 0L),
        end = fifelse(strand == "+", start + FLANK, end + FLANK),
        gene_name, Individual_ID, Haplotype, sample_hap, strand
    )]
}

revived_regions <- make_regions(revived_fiber)
dead_regions <- make_regions(dead_fiber)
positive_regions <- make_regions(positive_fiber)

peaks_bed <- peaks[, .(chrom, start, end, fire_cov_OV_cov)]
all_bed <- all_sites[, .(chrom, start, end, fire_cov_OV_cov)]

hits_for <- function(regions) {
    h <- bed_map(regions, peaks_bed,
        n_peaks = n(),
        max_peak_acc = ifelse(n() == 0, NA_real_, max(fire_cov_OV_cov, na.rm = TRUE))
    ) %>%
        replace_na(list(n_peaks = 0L)) %>%
        data.table()
    h[, max_acc := bed_map(
        regions, all_bed,
        max_acc = ifelse(n() == 0, NA_real_, max(fire_cov_OV_cov, na.rm = TRUE))
    )$max_acc]
    h
}

revived_hits <- hits_for(revived_regions)
dead_hits <- hits_for(dead_regions)
positive_hits <- hits_for(positive_regions)

cat("Revived with peaks:", sum(revived_hits$n_peaks > 0), "/", nrow(revived_hits), "\n")
cat("Dead with peaks:", sum(dead_hits$n_peaks > 0), "/", nrow(dead_hits), "\n")
cat("Positive control with peaks:", sum(positive_hits$n_peaks > 0), "/", nrow(positive_hits), "\n")
```

# Compare peak rates
```{r compare-peak-rates}
# subsample dead and positive-control loci to match revived count
set.seed(42)
n_revived <- nrow(revived_hits)
dead_hits_sub <- dead_hits[sample(.N, min(.N, n_revived))]
positive_hits_sub <- positive_hits[sample(.N, min(.N, n_revived))]
cat(
    "Subsampled dead loci:", nrow(dead_hits_sub),
    " positive-control loci:", nrow(positive_hits_sub),
    " to match", n_revived, "revived\n"
)

# fixed ordering: dead -> revived -> positive control (low -> high accessibility)
type_levels <- c("Dead pseudogene", "Revived pseudogene", "Protein-coding")

comparison <- rbind(
    revived_hits[, .(type = "Revived pseudogene", has_peak = n_peaks > 0, max_acc)],
    dead_hits_sub[, .(type = "Dead pseudogene", has_peak = n_peaks > 0, max_acc)],
    positive_hits_sub[, .(type = "Protein-coding", has_peak = n_peaks > 0, max_acc)]
)
comparison[, type := factor(type, levels = type_levels)]

print(comparison[, .(
    n = .N,
    n_with_peak = sum(has_peak),
    frac_with_peak = mean(has_peak)
), by = type])

# Fisher's test: enrichment of peaks in revived vs dead (kept as the primary contrast)
contingency <- matrix(c(
    sum(revived_hits$n_peaks > 0),
    sum(revived_hits$n_peaks == 0),
    sum(dead_hits_sub$n_peaks > 0),
    sum(dead_hits_sub$n_peaks == 0)
), nrow = 2, dimnames = list(
    c("With peak", "No peak"),
    c("Revived", "Dead")
))
cat("Contingency table (Revived vs Dead):\n")
print(contingency)

ft <- fisher.test(contingency)
cat("\nFisher's exact test (Revived vs Dead):\n")
cat("  Odds ratio:", round(ft$estimate, 2), "\n")
cat("  p-value:", ft$p.value, "\n")
cat("  95% CI:", round(ft$conf.int, 2), "\n")

# All pairwise Fisher tests for the enrichment plot brackets.
# fisher.test OR = odds(yes|col1) / odds(yes|col2), so the FIRST argument is
# the numerator. Convention here: more-active class first, so OR > 1 always
# means "the more-active class is enriched for peaks vs the reference".
peak_counts <- list(
    "Dead pseudogene"    = c(yes = sum(dead_hits_sub$n_peaks > 0), no = sum(dead_hits_sub$n_peaks == 0)),
    "Revived pseudogene" = c(yes = sum(revived_hits$n_peaks > 0), no = sum(revived_hits$n_peaks == 0)),
    "Protein-coding"     = c(yes = sum(positive_hits_sub$n_peaks > 0), no = sum(positive_hits_sub$n_peaks == 0))
)
fisher_pair <- function(numerator, denominator) {
    m <- matrix(c(peak_counts[[numerator]], peak_counts[[denominator]]),
        nrow = 2,
        dimnames = list(c("With peak", "No peak"), c(numerator, denominator))
    )
    fisher.test(m)
}

# numerator first: more-accessible class on the left
ft_rd <- fisher_pair("Revived pseudogene", "Dead pseudogene")
ft_pr <- fisher_pair("Protein-coding", "Revived pseudogene")
ft_pd <- fisher_pair("Protein-coding", "Dead pseudogene")

pair_labels <- c(
    ft_rd = "Revived vs Dead",
    ft_pr = "Protein-coding vs Revived",
    ft_pd = "Protein-coding vs Dead"
)
for (nm in names(pair_labels)) {
    f <- get(nm)
    cat("\nFisher's exact test (", pair_labels[[nm]], "):\n", sep = "")
    cat("  Odds ratio:", round(f$estimate, 2), "\n")
    cat("  p-value:", signif(f$p.value, 3), "\n")
}

p_bar <- comparison %>%
    ggplot(aes(x = type, fill = has_peak)) +
    geom_bar(position = "fill") +
    scale_y_continuous(
        "Fraction of promoters with Fiber-seq",
        labels = percent
    ) +
    scale_fill_manual("",
        values = c("TRUE" = "steelblue", "FALSE" = "grey80"),
        labels = c("TRUE" = "Peak nearby", "FALSE" = "No peak")
    ) +
    xlab("") +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/pseduo-genes-peak-rate.pdf", plot = p_bar, width = 3, height = 2)
```

# Accessibility distributions (all genes, no peak = 0)
```{r accessibility}
comparison[, acc_plot := fifelse(is.na(max_acc), 0, max_acc)]

p_acc <- comparison %>%
    ggplot(aes(x = acc_plot, color = type)) +
    stat_bin(geom = "step", binwidth = 0.03, position = "identity") +
    scale_x_continuous(
        "% Fiber-seq actuation near gene",
        labels = percent
    ) +
    scale_y_continuous("# of genes") +
    scale_color_brewer("", palette = "Set1") +
    facet_zoom(
        xlim = c(0.2, 1),
        ylim = c(0, 500),
        zoom.size = 2,
        horizontal = FALSE
    ) +
    theme_cowplot(
        font_size = 6,
    ) +
    theme(
        legend.position = "top"
    )
my_ggsave("Figures/pseduo-genes-accessibility.pdf", plot = p_acc, width = 3, height = 3)
```

# Combined figure
```{r combined-figure}
p_bar + p_acc + plot_layout(design = "ABB") +
    plot_annotation(tag_levels = "A")
my_ggsave("Figures/pseduo-genes-combined.pdf", width = 5, height = 3)
```

**Figure legend.** Pseudogenes and a protein-coding positive control identified across HPRC assemblies were intersected with fiber-seq accessibility peaks from 78 haplotype samples. **(A)** Fraction of revived pseudogene, dead pseudogene, and protein-coding loci (gene body +/- 500bp) overlapping at least one accessibility peak. **(B)** Distribution of maximum accessibility observed near each locus; bottom panel zooms into the 20-100% range.


# Alternate combined: enrichment bar + high-accessibility distribution
```{r alt-combined}
# semantic palette: revived = the biological finding (red), controls muted.
# Dead = negative control (neutral gray), Protein-coding = positive control
# (blue-gray — distinct from dead but still de-emphasized vs. revived).
pseudo_colors <- c(
    "Dead pseudogene"    = "#9E9E9E", # neutral gray (negative control)
    "Revived pseudogene" = "#C0392B", # red (finding)
    "Protein-coding"     = "#5B7C99" # slate blue-gray (positive control)
)

# enrichment bar chart with odds ratio annotation
peak_rates <- comparison[, .(
    frac = mean(has_peak)
), by = type]

bracket_label <- function(f, num_short, denom_short) {
    paste0("OR=", round(f$estimate, 2), ", p=", signif(f$p.value, 3))
}
y_max <- max(peak_rates$frac)

p_enrich <- peak_rates %>%
    ggplot(aes(x = type, y = frac)) +
    geom_bar(aes(fill = type), stat = "identity", width = 0.6) +
    geom_text(aes(label = paste0(round(100 * frac, 1), "%")),
        hjust = -0.3, size = 2.5
    ) +
    # adjacent pairs first, then the outermost (Dead vs Protein-coding) on top
    geom_bracket(
        xmin = "Dead pseudogene", xmax = "Revived pseudogene",
        y.position = y_max * 1.25,
        label = bracket_label(ft_rd, "Revived", "Dead"),
        label.size = 2, tip.length = 0.04, coord.flip = TRUE
    ) +
    geom_bracket(
        xmin = "Revived pseudogene", xmax = "Protein-coding",
        y.position = y_max * 1.45,
        label = bracket_label(ft_pr, "PC", "Revived"),
        label.size = 2, tip.length = 0.04, coord.flip = TRUE
    ) +
    geom_bracket(
        xmin = "Dead pseudogene", xmax = "Protein-coding",
        y.position = y_max * 1.65,
        label = bracket_label(ft_pd, "PC", "Dead"),
        label.size = 2, tip.length = 0.04, coord.flip = TRUE
    ) +
    scale_y_continuous("% of TSSs with a Fiber-seq peak",
        labels = percent, expand = expansion(mult = c(0, 0.2))
    ) +
    scale_fill_manual("", values = pseudo_colors) +
    xlab("") +
    coord_flip() +
    my_grid() +
    theme(legend.position = "none")

# high-accessibility distribution (>= 50%)
lower_limit <- 0.4
p_high_acc <- comparison[!is.na(acc_plot) & acc_plot >= lower_limit] %>%
    ggplot(aes(x = acc_plot, color = type)) +
    stat_bin(geom = "step", binwidth = 0.05, position = "identity") +
    scale_x_continuous("% Fiber-seq actuation",
        labels = percent, limits = c(lower_limit, 1)
    ) +
    scale_y_continuous("# of genes") +
    scale_color_manual("", values = pseudo_colors) +
    my_grid() +
    theme(legend.position = "top")

p_enrich + p_high_acc +
    plot_layout(design = "
    AB") +
    plot_annotation(tag_levels = "A")
my_ggsave("Figures/pseduo-genes-alt-combined.pdf", width = 6, height = 2)
```

**Figure legend.** Accessibility at gene promoters (TSS +/- 500bp) in HPRC Fiber-seq data across three classes: dead pseudogenes (negative control), revived pseudogenes, and protein-coding genes (positive control). **(A)** Fraction of loci in each class with at least one accessibility peak at the TSS. Brackets give Fisher's exact test odds ratios and p-values for all three pairwise contrasts. Dead and protein-coding loci were subsampled to match the number of revived pseudogenes. **(B)** Distribution of maximum fiber-seq actuation for loci with >= 40% accessibility.