```{r}
source("Rscripts/utils.R")
```

```{r}
all_manifest <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQANHblpVkMcVMWUvYas6JQl0pqvgeiYDlWwG2xV9vQdihgTsdWqHfl0z5BKb1I8hXcQtH1WIhwTkII/pub?output=csv") %>%
    # filter(Individual_ID != "HG02275") %>%
    filter(!is.na(raw_bam), raw_bam != "") %>%
    filter(!is.na(dsa), dsa != "") %>%
    filter(ID != "HG002_PS01015_PacBio", ID != "GM12878_PM00001_PacBio") %>%
    data.table()
glimpse(all_manifest)
glimpse(tail(all_manifest))
all_manifest %>%
    as.tibble()
all_manifest$ID
colnames(all_manifest)

all_manifest %>%
    fwrite("../fiber-seq-pilot/online-manifest.tbl", row.names = FALSE, quote = FALSE, sep = "\t")

manifest <- all_manifest
glimpse(manifest)
```

# load post-FIRE-filter QC tables (one row table per sample)
```{r}
read_qc <- function(x) {
    z <- fread(x)
    z$qc <- x
    z
}
# Vollger lab Redwood path; falls back to Stergachis lab path if running there.
qc_post_fire_files <- Sys.glob(
    "/scratch/ucgd/lustre-labs/vollger/projects/HPRCv2/fiber-seq-pilot/DSA-phasing/results/*.fire.filter.qc.tbl.gz"
)
if (length(qc_post_fire_files) == 0) {
    qc_post_fire_files <- Sys.glob(
        "/mmfs1/gscratch/stergachislab/HPRC/fiber-seq-pilot/DSA-FIRE/results/*/*qc.tbl.gz"
    )
}
qc_post_fire_files <- sort(qc_post_fire_files)

qc_fire_raw <- rbindlist(lapply(qc_post_fire_files, read_qc), idcol = "file")
qc_fire_raw[, uID := str_remove(basename(qc), "[.-].*")]
qc_fire_raw[, c("qc", "file") := NULL]

# Build a small per-sample lookup of (uID -> Individual_ID, PS, Platform)
# from the unique uIDs, then join it onto the big qc table. Doing the
# separate() on unique uIDs (~44 rows) instead of all ~70M rows is
# orders of magnitude faster.
# Drops the aggregate HG002_PS01015_PacBio (pilot-depth) because the
# HG002 deep aggregate is added back as its own group.
sample_lookup <- data.table(uID = unique(qc_fire_raw$uID))[
    uID != "HG002_PS01015_PacBio"
][, `:=`(
    Individual_ID = fifelse(
        uID == "HG002", "HG002",
        str_extract(uID, "^[^_]+")
    ),
    PS = fifelse(uID == "HG002", "HG002", str_extract(uID, "PS[0-9]+")),
    # Platform = sequencing technology. HG002 deep aggregate is PacBio.
    Platform = fcase(
        uID == "HG002", "PacBio",
        grepl("_PacBio$", uID), "PacBio",
        grepl("_ONT$", uID), "ONT"
    ),
    # group = reporting bucket. HG002 (added-on extra deep sample) is
    # tracked separately from the pilot PacBio/ONT samples.
    group = fcase(
        uID == "HG002", "HG002",
        grepl("_PacBio$", uID), "PacBio",
        grepl("_ONT$", uID), "ONT"
    )
)][!is.na(Platform)]

setkey(qc_fire_raw, uID)
setkey(sample_lookup, uID)
qc_fire_df <- qc_fire_raw[sample_lookup, nomatch = 0L]
qc_fire_df
```

# helpers used below
```{r}
n50_fun <- function(lengths, counts) {
    if (missing(counts)) counts <- rep(1, length(lengths))
    tdf <- data.frame(lengths = lengths, counts = counts) %>%
        arrange(lengths) %>%
        mutate(cum_counts = cumsum(counts * lengths)) %>%
        summarise(
            total_counts = sum(counts * lengths, na.rm = TRUE),
            n50 = lengths[cum_counts >= total_counts / 2][1]
        )
    tdf$n50[[1]]
}
```

# Build per-sample summary used by all paragraph chunks below.
# Total Gbp = sum of `phased_bp` (H1+H2+UNK) per sample.
# N50      = N50 of `fiber_length` distribution per sample.
```{r}
sample_summary <- qc_fire_df %>%
    filter(statistic == "phased_bp") %>%
    group_by(uID, Individual_ID, PS, Platform, group) %>%
    summarise(
        total_Gbp = sum(as.numeric(count), na.rm = TRUE) / 1e9,
        .groups = "drop"
    ) %>%
    left_join(
        qc_fire_df %>%
            filter(statistic == "fiber_length") %>%
            mutate(value = as.numeric(value)) %>%
            group_by(uID, Individual_ID, PS, Platform, group) %>%
            summarise(
                n50 = n50_fun(value, as.numeric(count)),
                .groups = "drop"
            ),
        by = c("uID", "Individual_ID", "PS", "Platform", "group")
    )

fwrite(sample_summary, "Tables/QC-per-sample-summary.tsv", sep = "\t")
```

# ---------------------------------------------------------------
# Q1: "For *** of the assemblies, we also generated long-read
# chromatin epigenomic data..."  -> count of unique individuals.
# ---------------------------------------------------------------
```{r}
n_indiv <- n_distinct(sample_summary$Individual_ID)
cat("Q1 — assemblies with Fiber-seq:", n_indiv, "\n")
```

# ---------------------------------------------------------------
# Q2 & Q3: "*** of the HPRC Fiber-seq-treated samples were
# subjected to PacBio sequencing (median Gbp ***, N50 ***), ***
# to ONT sequencing (median Gbp ***, N50 ***)"
# ---------------------------------------------------------------
```{r}
platform_summary <- sample_summary %>%
    group_by(group) %>%
    summarise(
        n_samples = n(),
        n_indiv = n_distinct(Individual_ID),
        median_Gbp = median(total_Gbp, na.rm = TRUE),
        median_N50 = median(n50, na.rm = TRUE),
        .groups = "drop"
    )
print(platform_summary)
fwrite(platform_summary, "Tables/QC-platform-summary.tsv", sep = "\t")
```

# ---------------------------------------------------------------
# Q4: "*** to both PacBio and ONT sequencing"
# ---------------------------------------------------------------
```{r}
both_platforms <- sample_summary %>%
    filter(group %in% c("PacBio", "ONT")) %>%
    distinct(Individual_ID, Platform) %>%
    count(Individual_ID) %>%
    filter(n == 2) %>%
    pull(Individual_ID)
cat(
    "Q4 — individuals with both PacBio and ONT:",
    length(both_platforms), "\n"
)
cat("    ", paste(both_platforms, collapse = ", "), "\n")
```

# ---------------------------------------------------------------
# Final printout: the four paragraph blanks we can fill from the
# per-sample QC tables.
# ---------------------------------------------------------------
```{r}
cat("=== PARAGRAPH FILL-IN VALUES ===\n\n")
cat(sprintf("Q1. Assemblies with Fiber-seq: %d\n", n_indiv))
for (g in c("PacBio", "ONT", "HG002")) {
    s <- platform_summary %>% filter(group == g)
    if (nrow(s) == 0) next
    cat(sprintf(
        "Q2/Q3. %s samples: %d (median Gbp = %.1f, N50 = %s bp)\n",
        g, s$n_samples, s$median_Gbp,
        scales::comma(round(s$median_N50))
    ))
}
cat(sprintf(
    "Q4. Both PacBio and ONT: %d  (%s)\n",
    length(both_platforms),
    paste(both_platforms, collapse = ", ")
))
```

# ---------------------------------------------------------------
# Q5 (figure): m6A autocorrelation by platform — supports the
# paragraph's claim that ONT shows reduced footprinting resolution
# vs PacBio while keeping a similar overall stenciling pattern.
# ---------------------------------------------------------------
```{r}
qc_fire_df %>%
    filter(Platform %in% c("PacBio", "ONT")) %>%
    filter(statistic == "m6a_acf") %>%
    mutate(value = as.numeric(value)) %>%
    filter(value > 30) %>%
    ggplot(aes(
        x = value, y = count,
        color = Individual_ID, group = uID
    )) +
    geom_line(linewidth = 0.3, alpha = 0.85) +
    scale_x_continuous("Lag between m6A sites (bp)") +
    scale_y_continuous("m6A autocorrelation") +
    my_grid() +
    facet_col(~Platform, scales = "free_y") +
    theme(legend.position = "none")
my_ggsave("Figures/m6a_acf_fire.pdf", width = 2.5, height = 2.5)
```

# ---------------------------------------------------------------
# Paragraph blanks NOT answered here (data lives in other scripts):
#   - peaks per haplotype (mean / range):
#       per-sample FIRE peak BEDs in DSA-FIRE/results/*/*peaks.bed.gz
#       see Rscripts/all-peaks.Rmd
#   - consensus pangenome peak count and # samples:
#       Tables/All-DSG-peaks.tsv.gz; see Rscripts/merged-DSA-peaks.Rmd
#   - non-GRCh38 peak count and percent:
#       GRCh38-beds/all-38.bed.gz; see Rscripts/compare-to-38.Rmd
#   - non-reference peak population frequency / SD enrichment:
#       same consensus table + annotations
# ---------------------------------------------------------------


# ===============================================================
# Legacy / unused plots — kept for reference, not required for
# answering the paragraph blanks above.
# These all use post-FIRE QC data sourced from Redwood paths.
# ===============================================================

# legacy: per-sample N50 bar plot
```{r}
sample_summary %>%
    filter(Platform %in% c("PacBio", "ONT")) %>%
    ggplot(aes(x = fct_reorder(uID, n50), y = n50, fill = Platform)) +
    geom_col() +
    scale_y_continuous("N50 (bp)", labels = scales::comma) +
    scale_x_discrete(labels = function(x) gsub("_.*", "", x)) +
    xlab("") +
    my_grid() +
    scale_fill_manual(values = PLAT_COLORS) +
    facet_row(~Platform, scales = "free_x", space = "free") +
    themen(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 4),
        legend.position = "none",
    )
my_ggsave("Figures/QC-n50-per-sample.pdf", width = 3, height = 2)
```


# legacy: per-sample total Gbp bar plot (post-FIRE only)
```{r}
sample_summary %>%
    filter(Platform %in% c("PacBio", "ONT")) %>%
    ggplot(aes(x = fct_reorder(uID, total_Gbp), y = total_Gbp, fill = Platform)) +
    geom_col() +
    scale_y_continuous("Gbp", labels = scales::comma) +
    scale_x_discrete(labels = function(x) gsub("_.*", "", x)) +
    xlab("") +
    my_grid() +
    scale_fill_manual(values = PLAT_COLORS) +
    facet_row(~Platform, scales = "free_x", space = "free") +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 4),
        legend.position = "none",
    )
my_ggsave("Figures/QC-phased-bp-per-sample.pdf", width = 3, height = 2)
```
