---
params:
  use_test: false
  use_mini: false
---

```{r setup}
source("Rscripts/utils.R")

# Global thresholds for calling a site a "peak" in a given sample
MIN_FIRE_COV <- 4
MIN_FRAC_ACC <- 0.20

# Haplotype analysis thresholds
DIFF_THRESHOLD <- 0.25
MIN_HAP_COV <- 10

# Modes:
#   USE_MINI = TRUE: ~30s mini-test (chr16 only, 4 individuals)
#   USE_TEST = TRUE: ~5min test (4 individuals, all chroms)
#   Both FALSE: full pipeline
# When run via pixi: controlled by params$use_test / params$use_mini
USE_MINI <- if (exists("params") && !is.null(params$use_mini)) params$use_mini else FALSE
USE_TEST <- if (exists("params") && !is.null(params$use_test)) params$use_test else FALSE
OUT_PREFIX <- fcase(
    USE_MINI, "Tables/mini-",
    USE_TEST, "Tables/test-",
    default = "Tables/"
)

UNION_PEAKS_FILE <- "/mmfs1/gscratch/stergachislab/HPRC/fiber-seq-pilot/HLA_graphs/peak_calling_all_chrs/All_HPRC_peaks_NewGraph_w_HG002/All_HPRC_peaks_NewGraph_w_HG002_final_tbl_w_metadata.tsv"
UNION_PEAKS_FILE <- "/mmfs1/gscratch/stergachislab/HPRC/fiber-seq-pilot/HLA_graphs/peak_calling_all_chrs/All_HPRC_peaks_NewGraph_w_HG002_LengthParam1/All_HPRC_peaks_NewGraph_w_HG002_LengthParam1_final_tbl_w_metadata.tsv"
TEST_FILE <- "Tables/union-peaks-test-input.tsv.gz"
MINI_FILE <- "Tables/union-peaks-mini-test-input.tsv.gz"
```

# Build test dataset (run once, then use cached file)
```{r build-test-data}
if (USE_TEST && !file.exists(TEST_FILE)) {
    cat("Building test dataset from full file...\n")
    full_data <- fread(UNION_PEAKS_FILE)
    full_data[, ind_key := sub("_[12]$", "", sample_id)]

    # pick one PacBio, one ONT, and HG002
    pb_key <- full_data[grepl("_PacBio_", sample_id), unique(ind_key)][1]
    ont_key <- full_data[grepl("_ONT_", sample_id), unique(ind_key)][1]
    hg002_keys <- full_data[grepl("^HG002_", sample_id), unique(ind_key)]
    test_keys <- unique(c(pb_key, ont_key, hg002_keys))
    cat("Test ind_keys:", paste(test_keys, collapse = ", "), "\n")

    test_data <- full_data[ind_key %in% test_keys]
    cat("Test rows:", nrow(test_data), "\n")

    fwrite(test_data[, !"ind_key"], TEST_FILE, sep = "\t")
    cat("Saved", TEST_FILE, "\n")
    rm(full_data, test_data)
    gc()
}
```

# Load chromosome alias table and segmental duplications
```{r load-annotations}
# Chromosome alias: maps assembly contig names -> hg38 chromosome names
all_chroms <- fread(
    "/mmfs1/gscratch/stergachislab/HPRC/assemblies/allChromAlias.txt",
    header = FALSE, col.names = c("chrom", "chrom_hg38", "chrom_ncbi")
)

# Segmental duplications (union of assembly + T2T SDs)
SDs <- bind_rows(
    my_read_bed("/mmfs1/gscratch/stergachislab/HPRC/annotations/segdups/all-segdups-merged.bed.gz"),
    my_read_bed("/mmfs1/gscratch/stergachislab/HPRC/annotations/segdups/T2Tv2.0.SDs.bed.gz"),
)
```

# Load union peaks
```{r load-data}
input_file <- fcase(
    USE_MINI, MINI_FILE,
    USE_TEST, TEST_FILE,
    default = UNION_PEAKS_FILE
)
cat("Loading:", input_file, "\n")

raw_df <- fread(input_file) %>%
    # convert empty strings to NA for assembly/hg38 columns
    mutate(across(where(is.character), ~ {
        .x[.x == "" | .x == "NA_NA_NA"] <- NA_character_
        .x
    })) %>%
    mutate(score = na_if(score, -2)) %>%
    mutate(across(c(cons_start, cons_end, asm_start, asm_end, hg38_start, hg38_end), as.integer)) %>%
    mutate(
        is_peak = coalesce(fire_coverage >= MIN_FIRE_COV & fire_cov_OV_cov >= MIN_FRAC_ACC, FALSE),
    ) %>%
    data.table()

# Filter consensus peaks: keep only those that are a peak in at least one sample
passing_peaks <- raw_df[is_peak == TRUE, unique(consensus_peak_id)]
cat(
    "Total consensus peaks:", uniqueN(raw_df$consensus_peak_id), "\n",
    "Passing in >= 1 sample:", length(passing_peaks), "\n",
    "Filtered out:", uniqueN(raw_df$consensus_peak_id) - length(passing_peaks), "\n"
)
raw_df <- raw_df[consensus_peak_id %in% passing_peaks]
```

# Annotate: sample ID parsing, SD overlap, chromosome placement
```{r annotate}
# SD overlap on unique coordinates only (much faster than per-row)
cons_coords <- raw_df[, .(cons_chr, cons_start, cons_end)] %>%
    unique() %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    bed_map(SDs, is_SD_cons = n() > 0) %>%
    rename(cons_chr = chrom, cons_start = start, cons_end = end) %>%
    replace_na(list(is_SD_cons = FALSE))
cat("SD overlap on", nrow(cons_coords), "unique cons coords\n")

asm_coords <- raw_df[!is.na(asm_chr), .(asm_chr, asm_start, asm_end)] %>%
    unique() %>%
    rename(chrom = asm_chr, start = asm_start, end = asm_end) %>%
    bed_map(SDs, is_SD_asm = n() > 0) %>%
    rename(asm_chr = chrom, asm_start = start, asm_end = end) %>%
    replace_na(list(is_SD_asm = FALSE))
cat("SD overlap on", nrow(asm_coords), "unique asm coords\n")

union_df <- raw_df %>%
    mutate(
        # parse sample_id: handles both "ID_PS_Platform_ID_Hap" and "ID_ID_Hap"
        Individual_ID = sub("_.*", "", sample_id),
        Haplotype = sub(".*_", "", sample_id),
        Platform = case_when(
            grepl("_PacBio_", sample_id) ~ "PacBio",
            grepl("_ONT_", sample_id) ~ "ONT",
            grepl("^HG002_HG002", sample_id) ~ "PacBio",
            TRUE ~ NA_character_
        ),
        PS = ifelse(
            str_count(sample_id, "_") >= 4,
            sub("^[^_]+_([^_]+)_.*", "\\1", sample_id),
            NA_character_
        ),
        ind_key = sub("_[12]$", "", sample_id),
    ) %>%
    # join pre-computed SD overlaps
    left_join(cons_coords, by = c("cons_chr", "cons_start", "cons_end")) %>%
    left_join(asm_coords, by = c("asm_chr", "asm_start", "asm_end")) %>%
    replace_na(list(is_SD_cons = FALSE, is_SD_asm = FALSE)) %>%
    mutate(is_SD = is_SD_cons | is_SD_asm) %>%
    # join asm_chr to chromosome alias table for hg38 placement
    left_join(
        all_chroms[, .(chrom, chrom_hg38)],
        by = c("asm_chr" = "chrom")
    ) %>%
    mutate(
        autosome = case_when(
            grepl("chrX|chrY", asm_chr) | grepl("chrX|chrY", chrom_hg38) ~ "Sex chromosome",
            TRUE ~ "Autosomes"
        ),
        # many-to-one: multiple original asm peaks merged into one consensus peak
        is_complex = num_overlapping_orig_peaks > 1,
        # which assembly the consensus peak is on
        cons_assembly = case_when(
            grepl("^chr[0-9XY]+_MATERNAL", cons_chr) ~ "HG002_1",
            grepl("^chr[0-9XY]+_PATERNAL", cons_chr) ~ "HG002_2",
            grepl("^chr[0-9XY]+$", cons_chr) ~ "CHM13",
            TRUE ~ sub("#[^#]+$", "", cons_chr) %>% sub("#", "_", .)
        ),
    ) %>%
    data.table()

cat("Merged (many asm -> 1 cons):", sum(union_df$is_complex), "\n")

# pick primary sample per individual:
# - For most individuals: PacBio over ONT
# - For HG002: pick the deep dataset (HG002_HG002, no PS)
primary_keys <- union_df[, .(
    mean_cov = mean(coverage, na.rm = TRUE)
), by = .(Individual_ID, ind_key, Platform, PS)]

# HG002 special case: pick the one without PS (deep PacBio)
primary_keys[
    Individual_ID == "HG002",
    is_primary_sample := is.na(PS)
]

# everyone else: pick PacBio if available, otherwise the only one
primary_keys[Individual_ID != "HG002", is_primary_sample := {
    if (.N == 1) {
        TRUE
    } else {
        Platform == "PacBio"
    }
}, by = Individual_ID]

union_df <- merge(
    union_df,
    primary_keys[, .(ind_key, is_primary_sample)],
    by = "ind_key", all.x = TRUE
)
cat("Primary samples:\n")
primary_keys[is_primary_sample == TRUE, .N, by = Platform]

rm(raw_df, cons_coords, asm_coords, primary_keys)
gc()

glimpse(union_df)
```


# ============================================================
# Haplotype analysis: pivot wide, classify, merge back
# ============================================================

# Pivot to wide via self-join (avoids dcast NA coercion issues)
```{r hap-pivot-wide}
merge_cols <- c("ind_key", "consensus_peak_id")
keep_cols <- c(
    "Individual_ID", "PS", "Platform",
    "cons_chr", "cons_start", "cons_end"
)
val_cols <- c(
    "coverage", "fire_coverage", "fire_cov_OV_cov",
    "is_peak", "asm_chr", "chrom_hg38", "is_SD", "autosome",
    "is_complex", "seq"
)

h1 <- union_df[Haplotype == "1", c(merge_cols, keep_cols, val_cols), with = FALSE]
h2 <- union_df[Haplotype == "2", c(merge_cols, val_cols), with = FALSE]

# sanity check: no duplicate keys (ensures 1:1 merge)
stopifnot(h1[, .N, by = .(ind_key, consensus_peak_id)][N > 1, .N] == 0)
stopifnot(h2[, .N, by = .(ind_key, consensus_peak_id)][N > 1, .N] == 0)

setnames(h1, val_cols, paste0(val_cols, "_1"))
setnames(h2, val_cols, paste0(val_cols, "_2"))

hap_wide <- merge(h1, h2, by = merge_cols, all = TRUE)

# derive combined and haplotype metrics
hap_wide[, `:=`(
    is_SD = fcoalesce(is_SD_1, FALSE) | fcoalesce(is_SD_2, FALSE),
    is_complex = fcoalesce(is_complex_1, FALSE) | fcoalesce(is_complex_2, FALSE),
    autosome = fifelse(
        autosome_1 %in% "Sex chromosome" | autosome_2 %in% "Sex chromosome",
        "Sex chromosome", "Autosomes"
    ),
    diff = fire_cov_OV_cov_1 - fire_cov_OV_cov_2,
    pass_hap_coverage = coverage_1 >= MIN_HAP_COV & coverage_2 >= MIN_HAP_COV,
    peak_status = fcase(
        is_peak_1 & is_peak_2, "both",
        is_peak_1 & !is_peak_2, "H1_only",
        !is_peak_1 & is_peak_2, "H2_only",
        default = "neither"
    )
)]

# keep only peaks where at least one haplotype is a peak
hap_wide <- hap_wide[peak_status != "neither"]
cat("Rows in hap_wide:", nrow(hap_wide), "\n")

cat("peak_status:\n")
hap_wide[, .N, by = peak_status]
```

# Sequence comparison between haplotypes (Biostrings local alignment)
```{r seq-comparison}
library(Biostrings)
library(future.apply)
plan(multisession, workers = min(availableCores(), 16))

hap_wide[, `:=`(
    seq_len_1 = nchar(seq_1),
    seq_len_2 = nchar(seq_2)
)]
hap_wide[, `:=`(
    seq_min_len = pmin(seq_len_1, seq_len_2, na.rm = TRUE),
    seq_max_len = pmax(seq_len_1, seq_len_2, na.rm = TRUE)
)]
hap_wide[, sufficient_overlap := seq_min_len >= 0.75 * seq_max_len]
hap_wide[, seq_identical := seq_1 == seq_2]

# default values for identical / missing sequences
hap_wide[, `:=`(
    local_mismatches = NA_integer_,
    pct_seq_identity = NA_real_
)]
hap_wide[seq_identical == TRUE, `:=`(
    local_mismatches = 0L,
    pct_seq_identity = 1.0
)]

# Run local alignment only on non-identical pairs with both seqs present
to_align <- hap_wide[
    seq_identical == FALSE & !is.na(seq_1) & !is.na(seq_2),
    which = TRUE
]
cat("Aligning", length(to_align), "sequence pairs...\n")

if (length(to_align) > 0) {
    s1 <- DNAStringSet(hap_wide$seq_1[to_align])
    s2 <- DNAStringSet(hap_wide$seq_2[to_align])

    # batch in chunks for parallel processing and progress
    chunk_size <- 50000
    n_chunks <- ceiling(length(to_align) / chunk_size)
    cat("Processing in", n_chunks, "chunks of", chunk_size, "\n")

    aln_results <- future_lapply(seq_len(n_chunks), function(i) {
        idx <- ((i - 1) * chunk_size + 1):min(i * chunk_size, length(to_align))
        # try forward and reverse complement, take the better alignment
        aln_fwd <- pairwiseAlignment(s1[idx], s2[idx], type = "local")
        aln_rev <- pairwiseAlignment(
            s1[idx], reverseComplement(s2[idx]),
            type = "local"
        )
        use_rev <- score(aln_rev) > score(aln_fwd)
        nmis <- ifelse(use_rev, nmismatch(aln_rev), nmismatch(aln_fwd))
        pid_v <- ifelse(use_rev, pid(aln_rev), pid(aln_fwd))
        nmat <- ifelse(use_rev, nmatch(aln_rev), nmatch(aln_fwd))
        wd <- ifelse(use_rev, width(pattern(aln_rev)), width(pattern(aln_fwd)))
        list(
            nmismatch = nmis,
            pid = pid_v / 100,
            nmatch = nmat,
            width = wd,
            is_rev = use_rev
        )
    })

    nmis <- unlist(lapply(aln_results, `[[`, "nmismatch"))
    pid_v <- unlist(lapply(aln_results, `[[`, "pid"))
    nmatch_v <- unlist(lapply(aln_results, `[[`, "nmatch"))
    aln_w <- unlist(lapply(aln_results, `[[`, "width"))

    hap_wide[to_align, `:=`(
        local_mismatches = nmis,
        pct_seq_identity = pid_v,
        aln_match = nmatch_v,
        aln_width = aln_w
    )]

    rm(s1, s2, aln_results, nmis, pid_v, nmatch_v, aln_w)
    gc()
}

# has_het_variants: insufficient overlap OR alignment found mismatches
hap_wide[, has_het_variants := !sufficient_overlap |
    (!is.na(local_mismatches) & local_mismatches > 0)]

plan(sequential)

cat(
    "Het variant rate:",
    round(100 * mean(hap_wide$has_het_variants, na.rm = TRUE), 1), "%\n"
)
```

# Fisher's exact test for haplotype accessibility difference
```{r fisher-test}
library(future.apply)
plan(multisession, workers = min(availableCores(), 16))

hap_wide[, nah1 := coverage_1 - fire_coverage_1]
hap_wide[, nah2 := coverage_2 - fire_coverage_2]

cat("Running Fisher's tests on", nrow(hap_wide), "rows...\n")
hap_wide[, fisher_p := future_mapply(
    function(a, b, c, d) {
        if (any(is.na(c(a, b, c, d)))) {
            return(NA_real_)
        }
        fisher.test(matrix(c(a, b, c, d), nrow = 2))$p.value
    },
    fire_coverage_1, fire_coverage_2, nah1, nah2,
    future.seed = TRUE
)]
hap_wide[, c("nah1", "nah2") := NULL]
plan(sequential)

cat("Significant (p < 0.05):", sum(hap_wide$fisher_p < 0.05, na.rm = TRUE), "\n")

# Classify haplotype status (single pass, using both diff AND Fisher's test)
hap_wide[, haplotype_distinct := fcase(
    # sex chromosomes (check assembly hg38 mapping and consensus peak coords)
    grepl("chrX", chrom_hg38_1, fixed = TRUE) | grepl("chrX", chrom_hg38_2, fixed = TRUE) |
        grepl("^chrX", cons_chr, fixed = FALSE),
    "chrX",
    grepl("chrY", chrom_hg38_1, fixed = TRUE) | grepl("chrY", chrom_hg38_2, fixed = TRUE) |
        grepl("^chrY", cons_chr, fixed = FALSE),
    "chrY",
    # complex: multiple asm peaks merged into one consensus peak
    is_complex,
    "Complex",
    # genetically haplotype-specific: peak on one hap, no assembly on other
    peak_status == "H1_only" & is.na(asm_chr_2),
    "Genetically haplotype-specific",
    peak_status == "H2_only" & is.na(asm_chr_1),
    "Genetically haplotype-specific",
    # insufficient coverage to test haplotype selectivity
    !pass_hap_coverage,
    "Untestable",
    # haplotype-selective: diff AND fisher, with het variants
    abs(diff) >= DIFF_THRESHOLD & fisher_p <= 0.05 & has_het_variants == TRUE,
    "Haplotype-selective",
    # epigenetically haplotype-selective: diff AND fisher, no het variants
    abs(diff) >= DIFF_THRESHOLD & fisher_p <= 0.05 & has_het_variants == FALSE,
    "Epigenetically haplotype-selective",
    # shared: everything else with sufficient coverage
    default = "Shared"
)]

hap_wide[, .N, by = haplotype_distinct]
```

# Merge haplotype classification back to long format
```{r hap-merge-back}
# Extract test results (only populated for peaks that pass filters)
hap_class <- hap_wide[, .(
    ind_key, consensus_peak_id, haplotype_distinct, diff,
    has_het_variants, local_mismatches, pct_seq_identity,
    fisher_p
)]
rm(hap_wide)
gc()

# Compute raw coverage columns via self-join on ALL rows (not just peaks)
# These should always be populated when both haplotypes exist
cov_h1 <- union_df[
    Haplotype == "1",
    .(ind_key, consensus_peak_id,
        coverage_1 = coverage, fire_coverage_1 = fire_coverage
    )
]
cov_h2 <- union_df[
    Haplotype == "2",
    .(ind_key, consensus_peak_id,
        coverage_2 = coverage, fire_coverage_2 = fire_coverage
    )
]

# NA means no row for that haplotype (not measured / not applicable)
# 0 means measured with zero reads (preserved from original data)
raw_cov <- merge(cov_h1, cov_h2,
    by = c("ind_key", "consensus_peak_id"), all = TRUE
)
rm(cov_h1, cov_h2)

union_hap_df <- merge(
    union_df, hap_class,
    by = c("ind_key", "consensus_peak_id"),
    all.x = TRUE
)
union_hap_df <- merge(
    union_hap_df, raw_cov,
    by = c("ind_key", "consensus_peak_id"),
    all.x = TRUE
)
cat("Rows with haplotype class:", sum(!is.na(union_hap_df$haplotype_distinct)), "\n")
cat("Rows with coverage_1:", sum(!is.na(union_hap_df$coverage_1)), "\n")
cat("Rows with coverage_2:", sum(!is.na(union_hap_df$coverage_2)), "\n")

rm(union_df, hap_class, raw_cov)
gc()
```

# Save tables: three coordinate versions + hap_wide
```{r save-tables}
bgz_write <- function(dt, path) {
    # rename first col to #chrom for BED convention
    names(dt)[1] <- paste0("#", names(dt)[1])
    tmp <- sub("\\.gz$", "", path)
    fwrite(dt, tmp, sep = "\t", quote = FALSE, na = "NA")
    system(paste("bgzip -@ 8 -f", shQuote(tmp)))
    system(paste("tabix -@ 8 -p bed", shQuote(path)))
    cat("Saved", path, "(", nrow(dt), "rows )\n")
}

# consensus coords (all rows)
union_hap_df %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "union-peaks-cons.bed.gz"))

# consensus coords (peak-only)
union_hap_df %>%
    filter(is_peak) %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "peaks-cons.bed.gz"))
gc()

# CHM13 coords
union_hap_df %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    filter(grepl("^chr[0-9XY]+$", chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "union-peaks-chm13.bed.gz"))

union_hap_df %>%
    filter(is_peak) %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    filter(grepl("^chr[0-9XY]+$", chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "peaks-chm13.bed.gz"))
gc()

# assembly coords
union_hap_df %>%
    rename(chrom = asm_chr, start = asm_start, end = asm_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "union-peaks-asm.bed.gz"))

union_hap_df %>%
    filter(is_peak) %>%
    rename(chrom = asm_chr, start = asm_start, end = asm_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "peaks-asm.bed.gz"))
gc()

# hg38 coords
union_hap_df %>%
    rename(chrom = hg38_chr, start = hg38_start, end = hg38_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "union-peaks-hg38.bed.gz"))

union_hap_df %>%
    filter(is_peak) %>%
    rename(chrom = hg38_chr, start = hg38_start, end = hg38_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end) %>%
    bgz_write(paste0(OUT_PREFIX, "peaks-hg38.bed.gz"))
gc()

# Write README describing outputs and columns
readme <- glue('
# Union Peaks Tables

## Thresholds used
- MIN_FIRE_COV: {MIN_FIRE_COV} (minimum FIRE coverage to call a peak)
- MIN_FRAC_ACC: {MIN_FRAC_ACC} (minimum fraction accessibility to call a peak)
- DIFF_THRESHOLD: {DIFF_THRESHOLD} (minimum H1-H2 accessibility difference for haplotype-selective)
- MIN_HAP_COV: {MIN_HAP_COV} (minimum coverage on both haplotypes to test selectivity)

## Output files
All files are bgzipped BED files with a `#chrom` header line.
Each file contains the same columns but with different leading coordinates
(renamed to chrom/start/end), filtered to rows where those coordinates exist.

### All sites (every sample x consensus peak that passes in at least one sample)
- `union-peaks-cons.bed.gz`: All rows, leading coords are consensus (graph) coordinates
- `union-peaks-chm13.bed.gz`: Subset where consensus coords are bare chr* names (T2T-CHM13 reference paths)
- `union-peaks-asm.bed.gz`: Rows with assembly coordinates (not NA)
- `union-peaks-hg38.bed.gz`: Rows with hg38 liftover coordinates (not NA)

### Peak-only (is_peak == TRUE)
Filtered to rows where the consensus peak passes thresholds in that specific sample
(fire_coverage >= {MIN_FIRE_COV} AND fire_cov_OV_cov >= {MIN_FRAC_ACC}).
Each row represents a called peak in a specific sample/haplotype.
- `peaks-cons.bed.gz`: Peaks only, consensus coordinates
- `peaks-chm13.bed.gz`: Peaks only, CHM13 coordinates
- `peaks-asm.bed.gz`: Peaks only, assembly coordinates
- `peaks-hg38.bed.gz`: Peaks only, hg38 coordinates

## Columns (in output order)

| # | Column | Description |
|---|--------|-------------|
| 1 | chrom | Leading coordinate chromosome (renamed from cons_chr, asm_chr, or hg38_chr depending on file) |
| 2 | start | Leading coordinate start |
| 3 | end | Leading coordinate end |
| 4 | ind_key | Individual+platform key (sample_id without haplotype suffix) |
| 5 | consensus_peak_id | Unique ID for the consensus peak in the graph |
| 6 | sample_id | Full sample identifier (e.g. HG00438_PS00991_ONT_HG00438_1) |
| 7 | coverage | Total fiber-seq coverage at this site in this sample |
| 8 | fire_coverage | Number of fibers showing accessible chromatin (FIRE) |
| 9 | score | Peak score from the caller |
| 10 | fire_cov_OV_cov | Fraction of fibers that are accessible (fire_coverage / coverage) |
| 11 | asm | Assembly identifier |
| 12 | asm_chr | Assembly contig name (NA if no assembly mapping) |
| 13 | asm_start | Assembly start coordinate |
| 14 | asm_end | Assembly end coordinate |
| 15 | asm_peak_id | Peak ID in assembly coordinates |
| 16 | hg38_chr | hg38 chromosome (NA if no liftover) |
| 17 | hg38_start | hg38 start coordinate |
| 18 | hg38_end | hg38 end coordinate |
| 19 | exists_in_hg38 | Whether the peak has hg38 coordinates (logical) |
| 20 | overlaps_original_called_peak | Whether this site overlaps a peak from the original per-sample analysis (logical) |
| 21 | sample_id_peak_id_TEMP_COL | Temporary join column from peak calling pipeline |
| 22 | num_overlapping_orig_peaks | Number of original per-sample peaks overlapping this consensus peak |
| 23 | orig_peaks_overlapping_consensus | IDs of original peaks overlapping this consensus peak |
| 24 | seq | DNA sequence at the peak |
| 25 | is_peak | Whether this site passes peak thresholds (fire_coverage >= {MIN_FIRE_COV} AND fire_cov_OV_cov >= {MIN_FRAC_ACC}) |
| 26 | Individual_ID | Individual identifier (e.g. HG00438), parsed from sample_id |
| 27 | Haplotype | Haplotype number (1 or 2), parsed from sample_id |
| 28 | Platform | Sequencing platform (PacBio or ONT; HG002_HG002 samples are PacBio) |
| 29 | PS | Sample preparation identifier, parsed from sample_id (NA for HG002) |
| 30 | is_SD_cons | Whether the consensus coordinates overlap a segmental duplication |
| 31 | is_SD_asm | Whether the assembly coordinates overlap a segmental duplication |
| 32 | is_SD | Whether either coordinate set overlaps a segmental duplication |
| 33 | chrom_hg38 | hg38 chromosome name for the assembly contig (from chromosome alias table) |
| 34 | autosome | "Autosomes" or "Sex chromosome", based on chrom_hg38/asm_chr |
| 35 | is_complex | Multiple original asm peaks merged into this consensus peak |
| 36 | cons_assembly | Assembly of consensus peak (e.g. HG00438_1, HG002_1, HG002_2, CHM13) |
| 37 | haplotype_distinct | Haplotype classification (see below) |
| 38 | diff | Accessibility difference between haplotypes (fire_cov_OV_cov H1 - H2) |
| 39 | has_het_variants | Whether the peak sequences differ between haplotypes (logical) |
| 40 | local_mismatches | Number of substitutions in best local alignment of haplotype sequences |
| 41 | pct_seq_identity | Percent sequence identity between haplotype sequences (0-1) |
| 42 | fisher_p | Fisher exact test p-value for accessibility difference between haplotypes |
| 43 | fire_coverage_1 | FIRE coverage on haplotype 1 for this individual |
| 44 | fire_coverage_2 | FIRE coverage on haplotype 2 for this individual |
| 45 | coverage_1 | Total coverage on haplotype 1 for this individual |
| 46 | coverage_2 | Total coverage on haplotype 2 for this individual |
| 47 | is_primary_sample | Whether this is the preferred sample for this individual (PacBio > ONT; deep HG002 > shallow) |

### haplotype_distinct categories
- **chrX**: Peak on chrX (excluded from autosomal haplotype analysis)
- **chrY**: Peak on chrY (excluded from autosomal haplotype analysis)
- **Complex**: Multiple original assembly peaks merged into this consensus peak on either haplotype
- **Genetically haplotype-specific**: Peak on one haplotype, no assembly mapping on the other
- **Haplotype-selective**: |diff| >= {DIFF_THRESHOLD} AND Fisher p <= 0.05, with het variants between haplotype sequences
- **Epigenetically haplotype-selective**: |diff| >= {DIFF_THRESHOLD} AND Fisher p <= 0.05, with identical haplotype sequences
- **Shared**: Sufficient coverage but does not meet both diff and Fisher thresholds
- **Untestable**: Insufficient coverage on one or both haplotypes to determine
- **NA**: Peak not called on either haplotype for this individual
')

writeLines(readme, paste0(OUT_PREFIX, "union-peaks-README.md"))
cat("Saved Tables/union-peaks-README.md\n")
```
