```{r setup}
source("Rscripts/utils.R")

# Global thresholds for calling a site a "peak" in a given sample
MIN_FIRE_COV <- 4
MIN_FRAC_ACC <- 0.20

# Haplotype analysis thresholds
DIFF_THRESHOLD <- 0.25
MIN_HAP_COV <- 10

# TRUE = load test subset (2 ind_keys: 1 PacBio + 1 ONT), FALSE = full data
USE_TEST <- FALSE
USE_TEST <- TRUE

UNION_PEAKS_FILE <- "/mmfs1/gscratch/stergachislab/HPRC/fiber-seq-pilot/HLA_graphs/peak_calling_all_chrs/All_HPRC_peaks_NewGraph_w_HG002/All_HPRC_peaks_NewGraph_w_HG002_final_tbl_w_metadata.tsv"
TEST_FILE <- "Tables/union-peaks-test-input.tsv.gz"
```

# Build test dataset (run once, then use cached file)
```{r build-test-data}
if (USE_TEST && !file.exists(TEST_FILE)) {
    cat("Building test dataset from full file...\n")
    full_data <- fread(UNION_PEAKS_FILE)
    full_data[, ind_key := sub("_[12]$", "", sample_id)]

    # pick one PacBio, one ONT, and HG002
    pb_key <- full_data[grepl("_PacBio_", sample_id), unique(ind_key)][1]
    ont_key <- full_data[grepl("_ONT_", sample_id), unique(ind_key)][1]
    hg002_keys <- full_data[grepl("^HG002_", sample_id), unique(ind_key)]
    test_keys <- unique(c(pb_key, ont_key, hg002_keys))
    cat("Test ind_keys:", paste(test_keys, collapse = ", "), "\n")

    test_data <- full_data[ind_key %in% test_keys]
    cat("Test rows:", nrow(test_data), "\n")

    fwrite(test_data[, !"ind_key"], TEST_FILE, sep = "\t")
    cat("Saved", TEST_FILE, "\n")
    rm(full_data, test_data)
    gc()
}
```

# Load chromosome alias table and segmental duplications
```{r load-annotations}
# Chromosome alias: maps assembly contig names -> hg38 chromosome names
all_chroms <- fread(
    "/mmfs1/gscratch/stergachislab/HPRC/assemblies/allChromAlias.txt",
    header = FALSE, col.names = c("chrom", "chrom_hg38", "chrom_ncbi")
)

# Segmental duplications (union of assembly + T2T SDs)
SDs <- bind_rows(
    my_read_bed("/mmfs1/gscratch/stergachislab/HPRC/annotations/segdups/all-segdups-merged.bed.gz"),
    my_read_bed("/mmfs1/gscratch/stergachislab/HPRC/annotations/segdups/T2Tv2.0.SDs.bed.gz"),
)
```

# Load union peaks
```{r load-data}
input_file <- ifelse(USE_TEST, TEST_FILE, UNION_PEAKS_FILE)
cat("Loading:", input_file, "\n")

raw_df <- fread(input_file) %>%
    # convert empty strings to NA for assembly/hg38 columns
    mutate(across(where(is.character), ~ {
        .x[.x == "" | .x == "NA_NA_NA"] <- NA_character_
        .x
    })) %>%
    mutate(score = na_if(score, -2)) %>%
    mutate(across(c(cons_start, cons_end, asm_start, asm_end, hg38_start, hg38_end), as.integer)) %>%
    mutate(
        is_peak = coalesce(fire_coverage >= MIN_FIRE_COV & fire_cov_OV_cov >= MIN_FRAC_ACC, FALSE),
    ) %>%
    data.table()

# Filter consensus peaks: keep only those that are a peak in at least one sample
passing_peaks <- raw_df[is_peak == TRUE, unique(consensus_peak_id)]
cat(
    "Total consensus peaks:", uniqueN(raw_df$consensus_peak_id), "\n",
    "Passing in >= 1 sample:", length(passing_peaks), "\n",
    "Filtered out:", uniqueN(raw_df$consensus_peak_id) - length(passing_peaks), "\n"
)
raw_df <- raw_df[consensus_peak_id %in% passing_peaks]
```

# Annotate: sample ID parsing, SD overlap, chromosome placement
```{r annotate}
union_df <- raw_df %>%
    mutate(
        # parse sample_id: handles both "ID_PS_Platform_ID_Hap" and "ID_ID_Hap"
        Individual_ID = sub("_.*", "", sample_id),
        Haplotype = sub(".*_", "", sample_id),
        Platform = case_when(
            grepl("_PacBio_", sample_id) ~ "PacBio",
            grepl("_ONT_", sample_id) ~ "ONT",
            TRUE ~ NA_character_
        ),
        PS = ifelse(
            str_count(sample_id, "_") >= 4,
            sub("^[^_]+_([^_]+)_.*", "\\1", sample_id),
            NA_character_
        ),
        ind_key = sub("_[12]$", "", sample_id),
    ) %>%
    # SD overlap via consensus coords
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    bed_map(SDs, is_SD_cons = n() > 0) %>%
    rename(cons_chr = chrom, cons_start = start, cons_end = end) %>%
    # SD overlap via assembly coords
    rename(chrom = asm_chr, start = asm_start, end = asm_end) %>%
    bed_map(SDs, is_SD_asm = n() > 0) %>%
    rename(asm_chr = chrom, asm_start = start, asm_end = end) %>%
    replace_na(list(is_SD_cons = FALSE, is_SD_asm = FALSE)) %>%
    mutate(is_SD = is_SD_cons | is_SD_asm) %>%
    # join asm_chr to chromosome alias table for hg38 placement
    left_join(
        all_chroms[, .(chrom, chrom_hg38)],
        by = c("asm_chr" = "chrom")
    ) %>%
    mutate(
        autosome = case_when(
            grepl("chrX|chrY", asm_chr) | grepl("chrX|chrY", chrom_hg38) ~ "Sex chromosome",
            TRUE ~ "Autosomes"
        ),
    ) %>%
    data.table()

glimpse(union_df)
```

# QC: summary stats
```{r summary-stats}
cat("Samples:", uniqueN(union_df$sample_id), "\n")
cat("Individuals:", uniqueN(union_df$Individual_ID), "\n")
cat("Consensus peaks:", uniqueN(union_df$consensus_peak_id), "\n")

union_df %>%
    group_by(Platform) %>%
    summarise(
        n_samples = n_distinct(sample_id),
        n_individuals = n_distinct(Individual_ID),
    )
```

# QC: peaks per sample
```{r qc-peaks-per-sample}
union_df %>%
    group_by(sample_id) %>%
    summarise(n_peaks = sum(is_peak))
```

# ============================================================
# Haplotype analysis: pivot wide, classify, merge back
# ============================================================

# Pivot to wide via self-join (avoids dcast NA coercion issues)
```{r hap-pivot-wide}
merge_cols <- c("ind_key", "consensus_peak_id")
keep_cols <- c(
    "Individual_ID", "PS", "Platform",
    "cons_chr", "cons_start", "cons_end"
)
val_cols <- c(
    "coverage", "fire_coverage", "fire_cov_OV_cov",
    "is_peak", "asm_chr", "chrom_hg38", "is_SD", "autosome"
)

h1 <- union_df[Haplotype == "1", c(merge_cols, keep_cols, val_cols), with = FALSE]
h2 <- union_df[Haplotype == "2", c(merge_cols, val_cols), with = FALSE]
setnames(h1, val_cols, paste0(val_cols, "_1"))
setnames(h2, val_cols, paste0(val_cols, "_2"))

hap_wide <- merge(h1, h2, by = merge_cols, all = TRUE)

# derive combined and haplotype metrics
hap_wide[, `:=`(
    is_SD = fcoalesce(is_SD_1, FALSE) | fcoalesce(is_SD_2, FALSE),
    autosome = fifelse(
        autosome_1 %in% "Sex chromosome" | autosome_2 %in% "Sex chromosome",
        "Sex chromosome", "Autosomes"
    ),
    diff = fire_cov_OV_cov_1 - fire_cov_OV_cov_2,
    pass_hap_coverage = coverage_1 >= MIN_HAP_COV & coverage_2 >= MIN_HAP_COV,
    peak_status = fcase(
        is_peak_1 & is_peak_2, "both",
        is_peak_1 & !is_peak_2, "H1_only",
        !is_peak_1 & is_peak_2, "H2_only",
        default = "neither"
    )
)]

# keep only peaks where at least one haplotype is a peak
hap_wide <- hap_wide[peak_status != "neither"]
cat("Rows in hap_wide:", nrow(hap_wide), "\n")

# classify haplotype status
hap_wide[, haplotype_distinct := fcase(
    # genetically haplotype-specific: peak on one hap, no assembly on other
    peak_status == "H1_only" & is.na(asm_chr_2),
    "Genetically haplotype-specific",
    peak_status == "H2_only" & is.na(asm_chr_1),
    "Genetically haplotype-specific",
    # insufficient coverage to test haplotype selectivity
    !pass_hap_coverage,
    "Untestable",
    # epigenetically selective: large diff with sufficient coverage
    abs(diff) >= DIFF_THRESHOLD,
    "Epigenetically haplotype-selective",
    # shared: small diff with sufficient coverage
    default = "Shared"
)]

hap_wide[, .N, by = haplotype_distinct]
```

# Merge haplotype classification back to long format
```{r hap-merge-back}
hap_class <- hap_wide[, .(ind_key, consensus_peak_id, haplotype_distinct, diff)]

union_hap_df <- merge(
    union_df, hap_class,
    by = c("ind_key", "consensus_peak_id"),
    all.x = TRUE
)
cat("Rows with haplotype class:", sum(!is.na(union_hap_df$haplotype_distinct)), "\n")
```

# Save tables: three coordinate versions + hap_wide
```{r save-tables}
bgz_write <- function(dt, path) {
    # rename first col to #chrom for BED convention
    names(dt)[1] <- paste0("#", names(dt)[1])
    tmp <- sub("\\.gz$", "", path)
    fwrite(dt, tmp, sep = "\t", quote = FALSE)
    system(paste("bgzip -@ 4 -f", shQuote(tmp)))
    cat("Saved", path, "\n")
}

# consensus coords (all rows)
cons_df <- union_hap_df %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end)
bgz_write(cons_df, "Tables/union-peaks-cons.bed.gz")

# CHM13 coords (cons_chr with bare chr* names = CHM13)
chm13_df <- union_hap_df %>%
    rename(chrom = cons_chr, start = cons_start, end = cons_end) %>%
    filter(grepl("^chr[0-9XY]+$", chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end)
bgz_write(chm13_df, "Tables/union-peaks-chm13.bed.gz")

# assembly coords
asm_df <- union_hap_df %>%
    rename(chrom = asm_chr, start = asm_start, end = asm_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end)
bgz_write(asm_df, "Tables/union-peaks-asm.bed.gz")

# hg38 coords
hg38_df <- union_hap_df %>%
    rename(chrom = hg38_chr, start = hg38_start, end = hg38_end) %>%
    filter(!is.na(chrom)) %>%
    select(chrom, start, end, everything()) %>%
    arrange(chrom, start, end)
bgz_write(hg38_df, "Tables/union-peaks-hg38.bed.gz")

# Filtered versions: only rows with asm_peak_id (assembly-level peak calls)
bgz_write(
    cons_df %>% filter(!is.na(asm_peak_id)),
    "Tables/union-peaks-cons-with-asm-peak.bed.gz"
)
bgz_write(
    chm13_df %>% filter(!is.na(asm_peak_id)),
    "Tables/union-peaks-chm13-with-asm-peak.bed.gz"
)
bgz_write(
    asm_df %>% filter(!is.na(asm_peak_id)),
    "Tables/union-peaks-asm-with-asm-peak.bed.gz"
)
bgz_write(
    hg38_df %>% filter(!is.na(asm_peak_id)),
    "Tables/union-peaks-hg38-with-asm-peak.bed.gz"
)

# Write README describing outputs and columns
readme <- glue('
# Union Peaks Tables

## Thresholds used
- MIN_FIRE_COV: {MIN_FIRE_COV} (minimum FIRE coverage to call a peak)
- MIN_FRAC_ACC: {MIN_FRAC_ACC} (minimum fraction accessibility to call a peak)
- DIFF_THRESHOLD: {DIFF_THRESHOLD} (minimum H1-H2 accessibility difference for haplotype-selective)
- MIN_HAP_COV: {MIN_HAP_COV} (minimum coverage on both haplotypes to test selectivity)

## Output files
All files are bgzipped BED files with a `#chrom` header line.
Each file contains the same columns but with different leading coordinates
(renamed to chrom/start/end), filtered to rows where those coordinates exist.

- `union-peaks-cons.bed.gz`: All rows, leading coords are consensus (graph) coordinates
- `union-peaks-chm13.bed.gz`: Subset where consensus coords are bare chr* names (T2T-CHM13 reference paths)
- `union-peaks-asm.bed.gz`: Rows with assembly coordinates (not NA)
- `union-peaks-hg38.bed.gz`: Rows with hg38 liftover coordinates (not NA)

## Columns (in output order)

| # | Column | Description |
|---|--------|-------------|
| 1 | chrom | Leading coordinate chromosome (renamed from cons_chr, asm_chr, or hg38_chr depending on file) |
| 2 | start | Leading coordinate start |
| 3 | end | Leading coordinate end |
| 4 | ind_key | Individual+platform key (sample_id without haplotype suffix) |
| 5 | consensus_peak_id | Unique ID for the consensus peak in the graph |
| 6 | sample_id | Full sample identifier (e.g. HG00438_PS00991_ONT_HG00438_1) |
| 7 | coverage | Total fiber-seq coverage at this site in this sample |
| 8 | fire_coverage | Number of fibers showing accessible chromatin (FIRE) |
| 9 | score | Peak score from the caller |
| 10 | fire_cov_OV_cov | Fraction of fibers that are accessible (fire_coverage / coverage) |
| 11 | asm | Assembly identifier |
| 12 | asm_chr | Assembly contig name (NA if no assembly mapping) |
| 13 | asm_start | Assembly start coordinate |
| 14 | asm_end | Assembly end coordinate |
| 15 | asm_peak_id | Peak ID in assembly coordinates |
| 16 | hg38_chr | hg38 chromosome (NA if no liftover) |
| 17 | hg38_start | hg38 start coordinate |
| 18 | hg38_end | hg38 end coordinate |
| 19 | exists_in_hg38 | Whether the peak has hg38 coordinates (logical) |
| 20 | overlaps_original_called_peak | Whether this site overlaps a peak from the original per-sample analysis (logical) |
| 21 | sample_id_peak_id_TEMP_COL | Temporary join column from peak calling pipeline |
| 22 | num_overlapping_orig_peaks | Number of original per-sample peaks overlapping this consensus peak |
| 23 | orig_peaks_overlapping_consensus | IDs of original peaks overlapping this consensus peak |
| 24 | seq | DNA sequence at the peak |
| 25 | is_peak | Whether this site passes peak thresholds (fire_coverage >= {MIN_FIRE_COV} AND fire_cov_OV_cov >= {MIN_FRAC_ACC}) |
| 26 | Individual_ID | Individual identifier (e.g. HG00438), parsed from sample_id |
| 27 | Haplotype | Haplotype number (1 or 2), parsed from sample_id |
| 28 | Platform | Sequencing platform (PacBio, ONT, or NA for HG002) |
| 29 | PS | Sample preparation identifier, parsed from sample_id (NA for HG002) |
| 30 | is_SD_cons | Whether the consensus coordinates overlap a segmental duplication |
| 31 | is_SD_asm | Whether the assembly coordinates overlap a segmental duplication |
| 32 | is_SD | Whether either coordinate set overlaps a segmental duplication |
| 33 | chrom_hg38 | hg38 chromosome name for the assembly contig (from chromosome alias table) |
| 34 | autosome | "Autosomes" or "Sex chromosome", based on chrom_hg38/asm_chr |
| 35 | haplotype_distinct | Haplotype classification (see below) |
| 36 | diff | Accessibility difference between haplotypes (fire_cov_OV_cov H1 - H2) |

### haplotype_distinct categories
- **Genetically haplotype-specific**: Peak on one haplotype, no assembly mapping on the other
- **Epigenetically haplotype-selective**: Sufficient coverage (>= {MIN_HAP_COV}) and |diff| >= {DIFF_THRESHOLD}
- **Shared**: Sufficient coverage and |diff| < {DIFF_THRESHOLD}
- **Untestable**: Insufficient coverage on one or both haplotypes to determine
- **NA**: Peak not called on either haplotype for this individual
')

writeLines(readme, "Tables/union-peaks-README.md")
cat("Saved Tables/union-peaks-README.md\n")
```


```{r}
nrow(cons_df)
nrow(cons_df[!is.na(asm_peak_id)])
glimpse(cons_df)
```

