```{r setup}
source("Rscripts/utils.R")
# Column definitions: see Tables/union-peaks-README.md
```

# Load tables (produced by union-peaks.Rmd)
```{r load-tables}
union_df <- fread(cmd = "bgzip -dc -@ 8 Tables/union-peaks-cons.bed.gz")
peak_df <- union_df[is_peak == TRUE]
nrow(peak_df)
nrow(union_df)
```

# Sample overview
```{r sample-overview}
# Peak-level summary (is_peak == TRUE)
sample_summary <- peak_df[, .(
    n_cons_peaks = .N,
    mean_coverage = mean(as.double(coverage), na.rm = TRUE),
    median_coverage = median(as.double(coverage), na.rm = TRUE),
    mean_acc = mean(fire_cov_OV_cov, na.rm = TRUE)
), by = .(Individual_ID, Haplotype, Platform, PS, sample_id)]

# All-site summary (including non-peaks)
site_summary <- union_df[, .(
    n_total = .N,
    n_with_asm = sum(!is.na(asm_chr)),
    n_with_hg38 = sum(!is.na(hg38_chr)),
    frac_asm = mean(!is.na(asm_chr)),
    frac_hg38 = mean(!is.na(hg38_chr))
), by = .(Individual_ID, Haplotype, Platform, PS, sample_id)]

sample_summary
```

# Peaks per sample by haplotype
```{r union-peaks-qc-peaks-per-hap}
sample_summary %>%
    # filter(!is.na(asm_chr)) %>%
    ggplot(aes(x = as.factor(Haplotype), y = n_cons_peaks)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("# of peaks per haplotype", labels = comma) +
    scale_x_discrete("Haplotype") +
    # scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-peaks-per-hap.pdf", width = 2, height = 2)
```

# Peaks per haplotype by HDCA class
```{r union-peaks-qc-unique-res}
re_per_hap <- peak_df[, .(
    n_peaks = .N
), by = .(sample_id, ind_key, Individual_ID, Platform, Haplotype)]

re_per_hap %>%
    ggplot(aes(x = as.factor(Haplotype), y = n_peaks)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("# regulatory elements per haplotype", labels = comma) +
    scale_x_discrete("Haplotype") +
    my_grid()
my_ggsave("Figures/union-peaks-qc-unique-res.pdf", width = 2, height = 2)
```

# REs per haplotype by HDCA class
```{r union-peaks-qc-unique-res-by-class}
re_by_class <- peak_df[autosome == "Autosomes", .(
    n_res = .N
), by = .(sample_id, ind_key, Individual_ID, Platform, Haplotype, haplotype_distinct)]

re_by_class[, haplotype_distinct := factor(
    haplotype_distinct,
    levels = HDCA_ORDER
)]

# compute median % of testable REs per class (excluding Untestable from denominator)
testable_total <- re_by_class[
    !is.na(haplotype_distinct) & haplotype_distinct != "Untestable",
    .(testable_total = sum(n_res)),
    by = sample_id
]
re_pct <- merge(re_by_class, testable_total, by = "sample_id")[
    !is.na(haplotype_distinct) & haplotype_distinct != "Untestable",
    .(median_pct = median(100 * n_res / testable_total)),
    by = haplotype_distinct
][, label := paste0(round(median_pct, 1), "%")]

re_by_class %>%
    filter(!is.na(haplotype_distinct)) %>%
    ggplot(aes(
        y = haplotype_distinct, x = n_res,
        fill = haplotype_distinct, color = haplotype_distinct
    )) +
    geom_quasirandom(
        method = "pseudorandom", size = 1, shape = 21,
        alpha = 0.3, stroke = 0.3, groupOnX = FALSE
    ) +
    geom_violin(color = "black", fill = NA, scale = "width", width = 0.7) +
    geom_text(
        data = re_pct,
        aes(
            x = 0, y = haplotype_distinct, label = label,
            color = NULL, fill = NULL
        ),
        hjust = -0.1, size = 2
    ) +
    scale_x_continuous("# regulatory elements", labels = comma) +
    scale_fill_manual(values = alpha(HDCA_COLORS, 0.5)) +
    scale_color_manual(values = colorspace::darken(HDCA_COLORS, 0.3)) +
    ylab("") +
    my_grid() +
    theme(legend.position = "none")
my_ggsave("Figures/union-peaks-qc-unique-res-by-class.pdf",
    width = 3, height = 2
)
```

# H1 vs H2 accessibility for epigenetically haplotype-selective peaks
```{r union-peaks-qc-epi-hap-sel-scatter}
epi_sel <- peak_df[haplotype_distinct == "Epigenetically haplotype-selective"]

epi_wide <- merge(
    epi_sel[Haplotype == "1", .(ind_key, consensus_peak_id, acc_1 = fire_cov_OV_cov)],
    epi_sel[Haplotype == "2", .(ind_key, consensus_peak_id, acc_2 = fire_cov_OV_cov)],
    by = c("ind_key", "consensus_peak_id")
)

epi_wide[, more_acc := fifelse(acc_1 > acc_2, "H1 > H2", "H2 > H1")] %>%
    ggplot(aes(x = acc_1, y = acc_2)) +
    geom_hex(bins = 50) +
    geom_xsidefreqpoly(aes(color = more_acc), alpha = 0.75, binwidth = 0.03) +
    geom_ysidefreqpoly(aes(color = more_acc), alpha = 0.75, binwidth = 0.03) +
    scale_fill_distiller("",
        palette = "Spectral",
        trans = "log10"
    ) +
    scale_color_manual("", values = c("H1 > H2" = "steelblue", "H2 > H1" = "firebrick")) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous("H1 accessibility", labels = percent, limits = c(NA, 1)) +
    scale_y_continuous("H2 accessibility", labels = percent, limits = c(NA, 1)) +
    scale_xsidey_continuous(labels = NULL) +
    scale_ysidex_continuous(labels = NULL) +
    my_grid() +
    theme(
        legend.position = "right",
        ggside.panel.scale = 0.2,
        aspect.ratio = 1,
    )

my_ggsave("Figures/union-peaks-qc-epi-hap-sel-scatter.pdf",
    width = 3, height = 3
)
```

# Sanity check: rows per (ind_key, consensus_peak_id) — should always be 2
```{r union-peaks-qc-rows-per-peak}
rows_per_peak <- union_df[, .N, by = .(ind_key, consensus_peak_id)]
rows_per_peak[, .N, by = N]
```

# H1 vs H2 peak counts per individual
```{r union-peaks-qc-h1-vs-h2-peaks}
h1_h2 <- sample_summary[, .(
    n_cons_peaks_1 = n_cons_peaks[Haplotype == "1"],
    n_cons_peaks_2 = n_cons_peaks[Haplotype == "2"]
), by = .(Individual_ID, Platform, PS)]

h1_h2 %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = n_cons_peaks_1, y = n_cons_peaks_2, color = Platform)) +
    geom_point(alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous("H1 peaks", labels = comma) +
    scale_y_continuous("H2 peaks", labels = comma) +
    scale_color_manual(values = PLAT_COLORS) +
    coord_equal() +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-h1-vs-h2-peaks.pdf", width = 2, height = 2)
```

# Coverage distribution per sample
```{r union-peaks-qc-coverage-per-sample}
sample_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = mean_coverage, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Mean fiber-seq coverage") +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-coverage-per-sample.pdf",
    width = 2, height = 2
)
```

# Accessibility distribution (sampled)
```{r union-peaks-qc-accessibility-dist}
set.seed(42)
plot_sample <- peak_df[sample(.N, min(.N, 1e6))]

plot_sample %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = Platform)) +
    geom_density(alpha = 0.5) +
    scale_x_continuous("Accessibility (FIRE / total coverage)",
        limits = c(0, 1)
    ) +
    scale_fill_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-accessibility-dist.pdf",
    width = 3, height = 2
)
```

# Accessibility by haplotype
```{r union-peaks-qc-accessibility-by-hap}
plot_sample %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = Haplotype)) +
    geom_density(alpha = 0.5) +
    scale_x_continuous("Accessibility (FIRE / total coverage)",
        limits = c(0, 1)
    ) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-accessibility-by-hap.pdf",
    width = 3, height = 2
)
```

# Fraction with assembly mapping
```{r union-peaks-qc-asm-mapping-rate}
site_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_asm, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction with assembly mapping",
        labels = percent, limits = c(0, 1)
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-asm-mapping-rate.pdf",
    width = 2, height = 2
)
```

# Fraction with hg38 liftover
```{r union-peaks-qc-hg38-liftover-rate}
site_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_hg38, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction with hg38 liftover",
        labels = percent, limits = c(0, 1)
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-hg38-liftover-rate.pdf",
    width = 2, height = 2
)
```

# Original peak recovery
```{r union-peaks-qc-orig-peak-recovery}
peak_df[!is.na(Platform),
    .(frac_orig = mean(overlaps_original_called_peak == TRUE)),
    by = .(Individual_ID, Platform, Haplotype, sample_id)
] %>%
    ggplot(aes(x = "", y = frac_orig * 100, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("% peaks overlapping original call") +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-orig-peak-recovery.pdf",
    width = 2, height = 2
)
```

# SD overlap rate per sample
```{r union-peaks-qc-sd-rate}
peak_df[, .(
    frac_sd = mean(is_SD)
), by = .(Individual_ID, Platform, Haplotype, sample_id)] %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_sd, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction of peaks in segmental duplications",
        labels = percent
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-sd-rate.pdf", width = 2, height = 2)
```

# Autosome vs sex chromosome counts
```{r union-peaks-qc-autosome-vs-sex}
peak_df[, .(
    n_auto = sum(autosome == "Autosomes"),
    n_sex = sum(autosome == "Sex chromosome")
), by = .(Individual_ID, Platform, Haplotype, sample_id)] %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = n_auto, y = n_sex, color = Platform)) +
    geom_point(alpha = 0.7) +
    scale_x_continuous("Autosomal peaks", labels = comma) +
    scale_y_continuous("Sex chromosome peaks", labels = comma) +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-autosome-vs-sex.pdf",
    width = 2, height = 2
)
```

# Peak size distribution
```{r union-peaks-qc-peak-size}
set.seed(42)
peak_df[sample(.N, min(.N, 500000))] %>%
    mutate(size = end - start) %>%
    ggplot(aes(x = size)) +
    geom_density(fill = "steelblue", alpha = 0.5) +
    scale_x_continuous("Peak size (bp)",
        trans = "log10", labels = comma,
        limits = c(50, 5000)
    ) +
    annotation_logticks(sides = "b") +
    my_grid()
my_ggsave("Figures/union-peaks-qc-peak-size.pdf", width = 3, height = 2)
```

# Haplotype classification summary
```{r union-peaks-qc-hap-class-summary}
hap_counts <- union_df[
    autosome == "Autosomes" &
        !is.na(haplotype_distinct) & haplotype_distinct != "Untestable",
    .N,
    by = haplotype_distinct
][, `:=`(
    total = sum(N),
    frac = N / sum(N),
    label = paste0(comma(N), " (", round(100 * N / sum(N), 1), "%)")
)]
hap_counts[, haplotype_distinct := factor(
    haplotype_distinct,
    levels = HDCA_ORDER
)]

hap_counts %>%
    ggplot(aes(
        x = N, y = haplotype_distinct,
        fill = haplotype_distinct
    )) +
    geom_bar(stat = "identity") +
    geom_text(aes(x = 0, label = label),
        hjust = -0.05, size = 2, color = "black"
    ) +
    scale_fill_manual(values = HDCA_COLORS) +
    scale_x_continuous("# of regulatory elements", labels = comma) +
    ylab("") +
    my_grid() +
    theme(legend.position = "none")
my_ggsave("Figures/union-peaks-qc-hap-class-summary.pdf",
    width = 3, height = 2
)
```

# Accessibility difference by haplotype classification
```{r union-peaks-qc-diff-by-hdca}
union_df[autosome == "Autosomes" & !is.na(haplotype_distinct)] %>%
    mutate(
        # for genetically haplotype-specific, use the peak haplotype's accessibility as diff
        diff_plot = case_when(
            haplotype_distinct == "Genetically haplotype-specific" ~ fire_cov_OV_cov,
            TRUE ~ diff
        )
    ) %>%
    filter(!is.na(diff_plot)) %>%
    mutate(haplotype_distinct = factor(
        haplotype_distinct,
        levels = HDCA_ORDER
    )) %>%
    ggplot(aes(x = diff_plot, fill = haplotype_distinct)) +
    geom_density(alpha = 0.5, adjust = 1) +
    scale_x_continuous(
        "Accessibility difference (H1 - H2)",
        limits = c(-1, 1)
    ) +
    scale_fill_manual(values = HDCA_COLORS) +
    facet_wrap(~haplotype_distinct, scales = "free_y", ncol = 1) +
    my_grid() +
    theme(legend.position = "none")
my_ggsave("Figures/union-peaks-qc-diff-by-hdca.pdf",
    width = 3, height = 3
)
```
