```{r setup}
source("Rscripts/utils.R")
# Column definitions: see Tables/union-peaks-README.md
```

# Load tables (produced by union-peaks.Rmd)
```{r load-tables}
union_df <- my_read_bed(cmd = "bgzip -dc -@ 8 Tables/union-peaks-cons.bed.gz", sort = FALSE)
peak_df <- union_df[is_peak == TRUE]
nrow(peak_df)
nrow(union_df)
```

# Sample overview
```{r sample-overview}
# Peak-level summary (is_peak == TRUE)
sample_summary <- peak_df[, .(
    n_cons_peaks = .N,
    mean_coverage = mean(as.double(coverage), na.rm = TRUE),
    median_coverage = median(as.double(coverage), na.rm = TRUE),
    mean_acc = mean(fire_cov_OV_cov, na.rm = TRUE)
), by = .(Individual_ID, Haplotype, Platform, PS, sample_id)]

# All-site summary (including non-peaks)
site_summary <- union_df[, .(
    n_total = .N,
    n_with_asm = sum(!is.na(asm_chr)),
    n_with_hg38 = sum(!is.na(hg38_chr)),
    frac_asm = mean(!is.na(asm_chr)),
    frac_hg38 = mean(!is.na(hg38_chr))
), by = .(Individual_ID, Haplotype, Platform, PS, sample_id)]

sample_summary
```

# Peaks per sample by haplotype
```{r union-peaks-qc-peaks-per-hap}
sample_summary %>%
    # filter(!is.na(asm_chr)) %>%
    ggplot(aes(x = as.factor(Haplotype), y = n_cons_peaks)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("# of peaks per haplotype", labels = comma) +
    scale_x_discrete("Haplotype") +
    # scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-peaks-per-hap.pdf", width = 2, height = 2)
```

# Peaks per haplotype
```{r union-peaks-qc-unique-res}
re_per_hap <- peak_df[, .(
    n_peaks = .N
), by = .(sample_id, ind_key, Individual_ID, Platform, Haplotype)]

re_per_hap %>%
    ggplot(aes(x = as.factor(Haplotype), y = n_peaks)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("# regulatory elements per haplotype", labels = comma) +
    scale_x_discrete("Haplotype") +
    my_grid()
my_ggsave("Figures/union-peaks-qc-unique-res.pdf", width = 2, height = 2)
```

# H1 vs H2 peak counts per individual

# Sanity check: rows per (ind_key, consensus_peak_id) — should always be 2
```{r union-peaks-qc-rows-per-peak}
rows_per_peak <- union_df[, .N, by = .(ind_key, consensus_peak_id)]
rows_per_peak[, .N, by = N]
```

# H1 vs H2 peak counts per individual
```{r union-peaks-qc-h1-vs-h2-peaks}
h1_h2 <- sample_summary[, .(
    n_cons_peaks_1 = n_cons_peaks[Haplotype == "1"],
    n_cons_peaks_2 = n_cons_peaks[Haplotype == "2"]
), by = .(Individual_ID, Platform, PS)]

h1_h2 %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = n_cons_peaks_1, y = n_cons_peaks_2, color = Platform)) +
    geom_point(alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    scale_x_continuous("H1 peaks", labels = comma) +
    scale_y_continuous("H2 peaks", labels = comma) +
    scale_color_manual(values = PLAT_COLORS) +
    coord_equal() +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-h1-vs-h2-peaks.pdf", width = 2, height = 2)
```

# Coverage distribution per sample
```{r union-peaks-qc-coverage-per-sample}
sample_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = mean_coverage, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Mean fiber-seq coverage") +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-coverage-per-sample.pdf",
    width = 2, height = 2
)
```

# Accessibility distribution (sampled)
```{r union-peaks-qc-accessibility-dist}
set.seed(42)
plot_sample <- peak_df[sample(.N, min(.N, 1e6))]

plot_sample %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = Platform)) +
    geom_density(alpha = 0.5) +
    scale_x_continuous("Accessibility (FIRE / total coverage)",
        limits = c(0, 1)
    ) +
    scale_fill_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-accessibility-dist.pdf",
    width = 3, height = 2
)
```

# Accessibility by haplotype
```{r union-peaks-qc-accessibility-by-hap}
plot_sample %>%
    ggplot(aes(x = fire_cov_OV_cov, fill = Haplotype)) +
    geom_density(alpha = 0.5) +
    scale_x_continuous("Accessibility (FIRE / total coverage)",
        limits = c(0, 1)
    ) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-accessibility-by-hap.pdf",
    width = 3, height = 2
)
```

# Fraction with assembly mapping
```{r union-peaks-qc-asm-mapping-rate}
site_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_asm, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction with assembly mapping",
        labels = percent, limits = c(0, 1)
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-asm-mapping-rate.pdf",
    width = 2, height = 2
)
```

# Fraction with hg38 liftover
```{r union-peaks-qc-hg38-liftover-rate}
site_summary %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_hg38, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction with hg38 liftover",
        labels = percent, limits = c(0, 1)
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-hg38-liftover-rate.pdf",
    width = 2, height = 2
)
```

# Original peak recovery
```{r union-peaks-qc-orig-peak-recovery}
peak_df[!is.na(Platform),
    .(frac_orig = mean(overlaps_original_called_peak == TRUE)),
    by = .(Individual_ID, Platform, Haplotype, sample_id)
] %>%
    ggplot(aes(x = "", y = frac_orig * 100, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("% peaks overlapping original call") +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-orig-peak-recovery.pdf",
    width = 2, height = 2
)
```

# SD overlap rate per sample
```{r union-peaks-qc-sd-rate}
peak_df[, .(
    frac_sd = mean(is_SD)
), by = .(Individual_ID, Platform, Haplotype, sample_id)] %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = "", y = frac_sd, color = Platform)) +
    geom_quasirandom(method = "pseudorandom") +
    geom_violin(color = "black", fill = NA) +
    scale_y_continuous("Fraction of peaks in segmental duplications",
        labels = percent
    ) +
    xlab("") +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-sd-rate.pdf", width = 2, height = 2)
```

# Autosome vs sex chromosome counts
```{r union-peaks-qc-autosome-vs-sex}
peak_df[, .(
    n_auto = sum(autosome == "Autosomes"),
    n_sex = sum(autosome == "Sex chromosome")
), by = .(Individual_ID, Platform, Haplotype, sample_id)] %>%
    filter(!is.na(Platform)) %>%
    ggplot(aes(x = n_auto, y = n_sex, color = Platform)) +
    geom_point(alpha = 0.7) +
    scale_x_continuous("Autosomal peaks", labels = comma) +
    scale_y_continuous("Sex chromosome peaks", labels = comma) +
    scale_color_manual(values = PLAT_COLORS) +
    my_grid() +
    theme(legend.position = "top")
my_ggsave("Figures/union-peaks-qc-autosome-vs-sex.pdf",
    width = 2, height = 2
)
```

# Peak size distribution
```{r union-peaks-qc-peak-size}
set.seed(42)
peak_df[sample(.N, min(.N, 500000))] %>%
    mutate(size = end - start) %>%
    ggplot(aes(x = size)) +
    geom_density(fill = "steelblue", alpha = 0.5) +
    scale_x_continuous("Peak size (bp)",
        trans = "log10", labels = comma,
        limits = c(50, 5000)
    ) +
    annotation_logticks(sides = "b") +
    my_grid()
my_ggsave("Figures/union-peaks-qc-peak-size.pdf", width = 3, height = 2)
```
